---
title: "Final Main and Supplementary Figure Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

이 문서는 **decontam 완료 후 저장된 phyloseq 객체**
`./Phyloseq/phy_F270R240_260812_v4.rds`를 불러오는 단계부터 시작합니다.
원자료 import, ASV table 구성 및 decontam 과정은 포함하지 않습니다. 저장된 객체에서
control sample을 제외하고 rarefaction한 뒤, 통계 분석과 각 패널 생성 및 최종 Figure
저장까지 수행합니다.

# Packages and plotting helper functions

```{r packages-and-functions, message=FALSE, warning=FALSE}
project_library <- file.path(getwd(), "R_library")
if (dir.exists(project_library)) .libPaths(c(project_library, .libPaths()))
suppressPackageStartupMessages({
  library(phyloseq)
  library(vegan)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(patchwork)
  library(ComplexUpset)
  library(pheatmap)
  library(RColorBrewer)
  library(ggpubr)
})


# Functions used directly in the final manuscript figures

theme_manuscript <- function(base_size = 8) {
  theme_classic(base_size = base_size) +
    theme(
      text = element_text(colour = "black"),
      axis.title = element_text(size = 9, face = "bold", colour = "black"),
      axis.text = element_text(size = 8, colour = "black"),
      plot.title = element_text(size = 10, face = "bold", hjust = 0.5,
                                margin = margin(b = 4)),
      plot.subtitle = element_text(size = 9, face = "bold", hjust = 0.5,
                                   margin = margin(b = 3)),
      strip.background = element_blank(),
      strip.text = element_text(size = 9, face = "bold"),
      legend.title = element_text(size = 9, face = "bold"),
      legend.text = element_text(size = 8),
      panel.grid = element_blank(),
      plot.margin = margin(6, 6, 6, 6)
    )
}

alpha_plot <- function(data, x_var, fill_var, shape_var, site_title,
                       y_title = NULL, paired = TRUE, show_legend = FALSE) {
  p <- ggplot(
    data,
    aes(x = .data[[x_var]], y = Value,
        fill = .data[[fill_var]], shape = .data[[shape_var]])
  ) +
    geom_boxplot(
      aes(group = interaction(.data[[x_var]], Index)),
      width = 0.65, outlier.shape = NA, colour = "black", linewidth = 0.45
    ) +
    geom_point(
      position = position_jitter(width = 0.08),
      size = 1.7, colour = "black"
    ) +
    ggpubr::geom_pwc(
      method = "wilcox_test",
      method.args = list(paired = paired, exact = FALSE),
      label = "{p.adj.signif}",
      p.adjust.method = "BH",
      step.increase = 0.10,
      hide.ns = TRUE
    ) +
    facet_wrap(~Index, scales = "free_y", nrow = 1) +
    labs(
      x = NULL, y = y_title, title = site_title,
      fill = ifelse(fill_var == "subject_id", "Participant", "Sampling type"),
      shape = ifelse(shape_var == "subject_id", "Participant", "Sampling type")
    ) +
    theme_manuscript()

  if (!show_legend) p <- p + theme(legend.position = "none")
  p
}

beta_plot <- function(ps, site, distance_method, test_variable,
                      colour_variable, shape_variable,
                      strata_variable = NULL, site_title,
                      metric_title, show_legend = TRUE) {
  keep <- sample_names(ps)[as.character(sample_data(ps)$site_group) == site]
  ps_site <- prune_samples(keep, ps)
  ps_site <- prune_taxa(taxa_sums(ps_site) > 0, ps_site)

  distance_object <- phyloseq::distance(ps_site, method = distance_method)
  ordination <- ordinate(ps_site, method = "PCoA", distance = distance_method)
  explained <- ordination$values$Relative_eig[1:2] * 100

  coordinates <- data.frame(ordination$vectors[, 1:2], check.names = FALSE)
  coordinates$sample_name <- rownames(coordinates)
  metadata <- data.frame(sample_data(ps_site), check.names = FALSE)
  metadata$sample_name <- rownames(metadata)
  coordinates <- left_join(coordinates, metadata, by = "sample_name")
  coordinates <- left_join(coordinates, swab_key, by = "swab_type2")

  set.seed(42)
  model_formula <- reformulate(test_variable, response = "distance_object")
  if (is.null(strata_variable)) {
    permanova <- adonis2(
      model_formula, data = metadata, permutations = 9999
    )
  } else {
    permanova <- adonis2(
      model_formula, data = metadata, permutations = 9999,
      strata = metadata[[strata_variable]]
    )
  }

  r2 <- permanova$R2[1]
  p_value <- permanova$`Pr(>F)`[1]
  p_label <- ifelse(
    p_value < 0.001, "<0.001",
    format(round(p_value, 3), nsmall = 3)
  )

  p <- ggplot(
    coordinates,
    aes(
      x = Axis.1, y = Axis.2,
      colour = .data[[colour_variable]],
      shape = .data[[shape_variable]]
    )
  ) +
    geom_vline(xintercept = 0, colour = "grey80") +
    geom_hline(yintercept = 0, colour = "grey80") +
    geom_point(size = 2, alpha = 0.75) +
    stat_ellipse(aes(group = .data[[colour_variable]]), linewidth = 0.55) +
    labs(
      x = paste0("PCoA 1 (", round(explained[1], 1), "%)"),
      y = paste0("PCoA 2 (", round(explained[2], 1), "%)"),
      title = site_title,
      subtitle = metric_title,
      colour = ifelse(
        colour_variable == "subject_id", "Participant", "Sampling type"
      ),
      shape = ifelse(
        shape_variable == "subject_id", "Participant", "Sampling type"
      )
    ) +
    annotate(
      "text", x = -Inf, y = -Inf, hjust = -0.10, vjust = -0.25, size = 2.7,
      label = paste0(
        "PERMANOVA:\nR2 = ", round(r2, 3),
        ", p-value = ", p_label
      )
    ) +
    theme_manuscript() +
    theme(aspect.ratio = 1)

  if (!show_legend) p <- p + theme(legend.position = "none")

  list(
    plot = p,
    permanova = data.frame(
      site = site_title,
      distance = metric_title,
      variable = test_variable,
      R2 = r2,
      F = permanova$F[1],
      p_value = p_value
    ),
    coordinates = coordinates
  )
}

taxa_plot <- function(melt, taxa, tax_otu, x_axis, phylum_or = NULL) {
  # Original taxon grouping, ordering, and phylum-specific colour logic.
  tax_tab <- unique(melt[, c("OTU", "Phylum", taxa)])
  tax_tab2 <- tax_tab[tax_tab$OTU %in% tax_otu, ]
  tax_phylum <- unique(tax_tab2$Phylum)

  plot_data <- melt
  plot_data[!plot_data$Phylum %in% tax_phylum, "Phylum"] <- "Other"
  plot_data[!plot_data$Phylum %in% tax_phylum, taxa] <- "Other"

  if (taxa != "Species") {
    for (phylum in tax_phylum) {
      selected_taxa <- tax_tab2[tax_tab2$Phylum == phylum, taxa]
      plot_data[
        plot_data$Phylum == phylum & !plot_data[, taxa] %in% selected_taxa,
        taxa
      ] <- paste0(phylum, "_Other")
    }
    for (phylum in tax_phylum) {
      selected_taxa <- tax_tab2[tax_tab2$Phylum == phylum, taxa]
      for (taxon in selected_taxa) {
        plot_data[plot_data[, taxa] == taxon, taxa] <- paste0(phylum, "_", taxon)
      }
    }
  } else {
    for (phylum in tax_phylum) {
      selected_taxa <- tax_tab2[tax_tab2$Phylum == phylum, taxa]
      replace_rows <- plot_data$Phylum == phylum & !plot_data[, taxa] %in% selected_taxa
      plot_data[replace_rows, taxa] <- "Other"
      plot_data[replace_rows, "Phylum"] <- "Other"
    }
  }

  abundance_summary <- plot_data[
    plot_data$Phylum %in% tax_phylum,
    c("Abundance", "Phylum", taxa)
  ] %>%
    group_by(Phylum, !!rlang::sym(taxa)) %>%
    summarise(sum.Abundance = sum(Abundance), .groups = "drop") %>%
    arrange(-sum.Abundance) %>%
    ungroup() %>%
    as.data.frame()

  phylum_order <- unique(abundance_summary$Phylum)
  plot_data$Phylum <- factor(plot_data$Phylum, levels = c(sort(phylum_order), "Other"))
  taxon_order <- abundance_summary %>% arrange(Phylum) %>% pull(!!rlang::sym(taxa))
  plot_data[, taxa] <- factor(plot_data[, taxa], levels = c(taxon_order, "Other"))

  phylum_color_map <- c(
    Actinobacteria = "Reds", Actinobacteriota = "Reds", Actinomycetota = "Reds",
    Firmicutes = "Blues", Bacillota = "Blues", Firmicutes_A = "Blues",
    Firmicutes_B = "BuGn", Firmicutes_C = "PuBu", Firmicutes_D = "YlGnBu",
    Bacteroidetes = "Purples", Bacteroidota = "Purples",
    Proteobacteria = "Greens", Pseudomonadota = "Greens",
    Fusobacteria = "YlOrBr", Fusobacteriota = "YlOrBr",
    Ascomycota = "RdPu", Basidiomycota = "YlOrBr"
  )
  other_palettes <- c("BrBG", "Spectral", "BrBG", "PuOr")
  categories <- abundance_summary %>%
    arrange(-sum.Abundance) %>% arrange(Phylum) %>% count(Phylum, name = "n_taxa")
  unlisted_phyla <- setdiff(categories$Phylum, names(phylum_color_map))
  if (length(unlisted_phyla) > 0) {
    phylum_color_map <- c(
      phylum_color_map,
      setNames(rep(other_palettes, length.out = length(unlisted_phyla)), unlisted_phyla)
    )
  }

  colour_list <- vector("list", nrow(categories))
  names(colour_list) <- categories$Phylum
  for (phylum in categories$Phylum) {
    n_taxa <- categories$n_taxa[categories$Phylum == phylum]
    palette_name <- phylum_color_map[[phylum]]
    if (n_taxa == 1) {
      colour_list[[phylum]] <- rev(brewer.pal(9, palette_name)[5])
    } else if (n_taxa == 2) {
      colour_list[[phylum]] <- rev(brewer.pal(9, palette_name)[c(3, 7)])
    } else if (n_taxa <= 9) {
      colour_list[[phylum]] <- rev(brewer.pal(n_taxa, palette_name))
    } else {
      colour_list[[phylum]] <- colorRampPalette(rev(brewer.pal(9, palette_name)))(n_taxa)
    }
  }
  colour_code <- c(unlist(colour_list, use.names = FALSE), "#D3D3D3")

  if (!is.null(phylum_or)) {
    sample_order <- plot_data %>%
      group_by(!!rlang::sym(x_axis)) %>% data.frame() %>%
      mutate(Abundance = Abundance / sum(Abundance)) %>%
      filter(Phylum %in% phylum_or) %>%
      group_by(!!rlang::sym(x_axis)) %>%
      summarise(Abundance = sum(Abundance), .groups = "drop") %>%
      arrange(Abundance) %>% pull(!!rlang::sym(x_axis)) %>% as.character()
    plot_data[, x_axis] <- factor(plot_data[, x_axis], levels = sample_order)
  }

  p <- ggplot(
    plot_data,
    aes(x = !!rlang::sym(x_axis), y = Abundance, fill = !!rlang::sym(taxa))
  ) +
    geom_bar(stat = "identity", position = "fill") +
    labs(y = "Relative abundance") +
    scale_fill_manual(values = colour_code) +
    theme_manuscript() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.title = element_blank())

  list(
    plot = p,
    data = list(df = plot_data, summary_df = abundance_summary),
    color = colour_code
  )
}

prepare_upset_data <- function(ps, site = NULL, mean_abundance_cutoff = 0.001) {
  ps_genus <- tax_glom(ps, taxrank = "Genus", NArm = FALSE)
  if (!is.null(site)) {
    keep <- sample_names(ps_genus)[
      as.character(sample_data(ps_genus)$site_group) == site
    ]
    ps_genus <- prune_samples(keep, ps_genus)
  }
  ps_genus <- prune_taxa(taxa_sums(ps_genus) > 0, ps_genus)

  ps_relative <- transform_sample_counts(ps_genus, function(x) x / sum(x))
  mean_relative <- taxa_sums(ps_relative) / nsamples(ps_relative)
  keep_taxa <- taxa_names(ps_relative)[mean_relative >= mean_abundance_cutoff]
  ps_genus <- prune_taxa(keep_taxa, ps_genus)

  metadata <- data.frame(sample_data(ps_genus), check.names = FALSE)
  otu <- as(otu_table(ps_genus), "matrix")
  if (!taxa_are_rows(ps_genus)) otu <- t(otu)

  presence <- sapply(swab_key$swab_type2, function(system) {
    samples <- rownames(metadata)[metadata$swab_type2 == system]
    rowSums(otu[, samples, drop = FALSE] > 0) > 0
  })
  presence <- data.frame(presence, check.names = FALSE)
  colnames(presence) <- as.character(swab_key$Type)
  presence$taxon_id <- rownames(presence)
  presence$Genus <- as.character(tax_table(ps_genus)[presence$taxon_id, "Genus"])

  system_columns <- paste0("Type ", 1:5)
  shared_ids <- presence$taxon_id[
    rowSums(presence[, system_columns, drop = FALSE]) == 5
  ]
  total_reads <- sum(otu)
  shared_reads <- sum(otu[shared_ids, , drop = FALSE])

  list(
    data = presence,
    summary = data.frame(
      subset = ifelse(is.null(site), "Total", site),
      retained_genera = nrow(presence),
      shared_by_all_five_genera = length(shared_ids),
      mean_abundance_cutoff = mean_abundance_cutoff,
      total_reads = total_reads,
      reads_from_shared_genera = shared_reads,
      shared_genus_relative_abundance_pct = 100 * shared_reads / total_reads
    )
  )
}

upset_plot <- function(upset_data, title, unit_label = "genera") {
  shared_label <- sprintf(
    "%d shared genera (%.1f%% of reads)",
    upset_data$summary$shared_by_all_five_genera,
    upset_data$summary$shared_genus_relative_abundance_pct
  )

  p <- ComplexUpset::upset(
    upset_data$data,
    intersect = paste0("Type ", 1:5),
    name = "Sampling system",
    min_size = 0,
    sort_sets = FALSE,
    sort_intersections_by = "cardinality",
    width_ratio = 0.2,
    base_annotations = list(
      "Intersection size" = intersection_size(
        counts = TRUE,
        bar_number_threshold = 1,
        text_colors = c(on_background = "black", on_bar = "black"),
        text_mapping = aes(
          label = ifelse(
            `Type 1` & `Type 2` & `Type 3` & `Type 4` & `Type 5`,
            "",
            exclusive_intersection_size
          )
        ),
        # Keep every count horizontal and centred above its bar.
        text = list(size = 3.3, angle = 0, hjust = 0.5, vjust = -0.65)
      ) +
        geom_text(
          aes(
            y = exclusive_intersection_size,
            label = ifelse(
              `Type 1` & `Type 2` & `Type 3` & `Type 4` & `Type 5`,
              shared_label,
              NA_character_
            )
          ),
          stat = "unique", colour = "black", size = 3.3,
          hjust = 0, vjust = -0.65, nudge_x = -0.45, na.rm = TRUE
        ) +
        scale_y_continuous(expand = expansion(mult = c(0, 0.30)))
    )
  )

  p <- p +
    labs(title = paste0(
      title, " (n=", upset_data$summary$retained_genera, " ", unit_label, ")"
    )) +
    theme_bw(base_size = 8) +
    theme(plot.title = element_text(face = "bold"), panel.grid = element_blank())
  p$theme$axis.text.x <- element_text(angle = 45, hjust = 1, vjust = 1)
  p
}
```

# Import post-decontam phyloseq RDS and prepare analysis data

```{r data-preprocessing, message=FALSE, warning=FALSE}
# -----------------------------------------------------------------------------
# 1. Start from the saved post-decontam phyloseq object
# -----------------------------------------------------------------------------

# This is the post-decontam RDS used as the starting point in the original Rmd.
phy <- readRDS("./Phyloseq/phy_F270R240_260812_v4.rds")

# Remove controls, then remove singleton ASVs (total abundance = 1)
# across the 50 true samples, following the filtering sensitivity analysis.
phy_true_unfiltered <- subset_samples(phy, control_status == "sample") %>%
  prune_taxa(taxa_sums(.) > 0, .)

phy <- prune_taxa(taxa_sums(phy_true_unfiltered) > 1, phy_true_unfiltered)

filter_summary <- data.frame(
  analysis = "canonical_singleton",
  retention_rule = "taxa_sums > 1",
  samples = nsamples(phy),
  asvs_before_filter = ntaxa(phy_true_unfiltered),
  asvs_after_filter = ntaxa(phy),
  asvs_removed = ntaxa(phy_true_unfiltered) - ntaxa(phy),
  reads_before_filter = sum(sample_sums(phy_true_unfiltered)),
  reads_after_filter = sum(sample_sums(phy)),
  reads_removed = sum(sample_sums(phy_true_unfiltered)) - sum(sample_sums(phy)),
  reads_retained_pct = 100 * sum(sample_sums(phy)) / sum(sample_sums(phy_true_unfiltered))
)

# Rarefy the post-decontam sample object using the existing analysis settings.
set.seed(42)
phy_rar <- rarefy_even_depth(
  phy,
  sample.size = min(sample_sums(phy)),
  rngseed = 42,
  replace = FALSE,
  verbose = FALSE
)

project_root <- sub("[/\\\\]+$", "", getwd())
figure_dir <- file.path(project_root, "Figures")
table_dir <- file.path(project_root, "Tables")
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
write.csv(filter_summary, file.path(table_dir, "feature_filter_summary.csv"), row.names = FALSE)

swab_key <- data.frame(
  swab_type2 = c("Copan", "Puritan_gel", "Puritan_buffer", "Omni", "Eswab"),
  Type = factor(paste0("Type ", 1:5), levels = paste0("Type ", 1:5))
)

type_col <- setNames(brewer.pal(5, "Set2"), paste0("Type ", 1:5))
subject_col <- setNames(
  brewer.pal(5, "Set1"),
  sort(unique(as.character(sample_data(phy)$subject_id)))
)

meta_rar <- data.frame(sample_data(phy_rar), check.names = FALSE)
meta_rar$sample_name <- sample_names(phy_rar)
meta_rar <- left_join(meta_rar, swab_key, by = "swab_type2")

# -----------------------------------------------------------------------------
```

# Figure 1 — bacterial diversity across sampling systems

```{r figure-1, message=FALSE, warning=FALSE}
# 2. Figure 1: beta and alpha diversity across sampling systems
# -----------------------------------------------------------------------------

# 1. Data processing
alpha_data <- estimate_richness(
  phy_rar, measures = c("Shannon", "Observed")
) %>%
  rownames_to_column("sample_name") %>%
  mutate(sample_name = sub("^X", "", sample_name)) %>%
  left_join(meta_rar, by = "sample_name") %>%
  pivot_longer(
    c(Shannon, Observed),
    names_to = "Index",
    values_to = "Value"
  ) %>%
  mutate(
    Index = factor(
      Index,
      levels = c("Shannon", "Observed"),
      labels = c("Shannon index", "Observed ASVs")
    )
  )

# 2. Beta-diversity statistics and plots
f1_A_result <- beta_plot(
  phy_rar, "Antecubital", "unifrac", "swab_type2",
  "Type", "subject_id", "subject_id",
  "Antecubital fossa", "Unweighted UniFrac", FALSE
)
f1_B_result <- beta_plot(
  phy_rar, "Antecubital", "wunifrac", "swab_type2",
  "Type", "subject_id", "subject_id",
  "Antecubital fossa", "Weighted UniFrac", TRUE
)
f1_C_result <- beta_plot(
  phy_rar, "Forehead", "unifrac", "swab_type2",
  "Type", "subject_id", "subject_id",
  "Forehead", "Unweighted UniFrac", FALSE
)
f1_D_result <- beta_plot(
  phy_rar, "Forehead", "wunifrac", "swab_type2",
  "Type", "subject_id", "subject_id",
  "Forehead", "Weighted UniFrac", TRUE
)

f1_A <- f1_A_result$plot + scale_colour_manual(values = type_col, drop = FALSE)
f1_B <- f1_B_result$plot + scale_colour_manual(values = type_col, drop = FALSE)
f1_C <- f1_C_result$plot + scale_colour_manual(values = type_col, drop = FALSE)
f1_D <- f1_D_result$plot + scale_colour_manual(values = type_col, drop = FALSE)

Figure1_PERMANOVA <- bind_rows(
  f1_A_result$permanova,
  f1_B_result$permanova,
  f1_C_result$permanova,
  f1_D_result$permanova
)

# 3. Alpha-diversity plots
f1_E <- alpha_plot(
  filter(alpha_data, site_group == "Antecubital"),
  x_var = "Type", fill_var = "Type", shape_var = "subject_id",
  site_title = "Antecubital fossa", paired = TRUE
) + scale_fill_manual(values = type_col, drop = FALSE)

f1_F <- alpha_plot(
  filter(alpha_data, site_group == "Forehead"),
  x_var = "Type", fill_var = "Type", shape_var = "subject_id",
  site_title = "Forehead", paired = TRUE
) + scale_fill_manual(values = type_col, drop = FALSE)

# 4. Assemble and inspect
Figure1 <- (
  wrap_elements(full = (f1_A | f1_B)) /
    wrap_elements(full = (f1_C | f1_D)) /
    (f1_E | f1_F)
) +
  plot_layout(heights = c(1, 1, 0.62)) +
  plot_annotation(tag_levels = list(c("A", "B", "C", "D"))) &
  theme(plot.tag = element_text(size = 16, face = "bold"))

print(Figure1)
print(Figure1_PERMANOVA)

# 5. Save the inspected objects
ggsave(
  file.path(figure_dir, "Figure1_bacterial_diversity.png"),
  plot = Figure1, width = 9, height = 10.2, dpi = 600,
  bg = "white", limitsize = FALSE
)
ggsave(
  file.path(figure_dir, "Figure1_bacterial_diversity.tiff"),
  plot = Figure1, width = 9, height = 10.2, dpi = 600,
  compression = "lzw", bg = "white", limitsize = FALSE
)
write.csv(
  Figure1_PERMANOVA,
  file.path(table_dir, "Figure1_PERMANOVA.csv"),
  row.names = FALSE
)

```




# Figure 2 — bacterial community composition

```{r figure-2, message=FALSE, warning=FALSE}
# 3. Figure 2: bacterial community composition
# -----------------------------------------------------------------------------
# 1. Data processing
phy_genus <- tax_glom(phy, taxrank = "Genus", NArm = FALSE)
phy_genus_relative <- transform_sample_counts(phy_genus, function(x) x / sum(x))

mean_genus_abundance <- taxa_sums(phy_genus_relative) / nsamples(phy_genus_relative)
major_genus_ids <- names(mean_genus_abundance)[mean_genus_abundance > 0.01]

genus_tax_plot <- taxa_plot(
  melt = psmelt(phy_genus),
  taxa = "Genus",
  tax_otu = major_genus_ids,
  x_axis = "swab_type2",
  phylum_or = "Actinobacteriota"
)

genus_composition_data <- genus_tax_plot$data$df %>%
  left_join(swab_key, by = "swab_type2")

genus_relative_data <- psmelt(phy_genus_relative) %>%
  left_join(swab_key, by = "swab_type2") %>%
  mutate(Genus = as.character(Genus))

selected_skin_genera <- c("Cutibacterium", "Staphylococcus", "Streptococcus")
selected_genus_data <- genus_relative_data %>%
  filter(Genus %in% selected_skin_genera)

# 2. Visualisation
f2_A <- ggplot(
  filter(genus_composition_data, site_group == "Antecubital"),
  aes(Type, Abundance, fill = Genus)
) +
  geom_col(position = "fill", width = 0.88) +
  scale_fill_manual(values = genus_tax_plot$color) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), expand = c(0, 0)) +
  labs(x = NULL, y = "Antecubital relative abundance (%)", fill = "Genus", tag = "A") +
  guides(fill = guide_legend(keyheight = unit(4.8, "mm"), keywidth = unit(7, "mm"))) +
  theme_manuscript() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        legend.text = element_text(size = 6),
        legend.key.height = unit(4.8, "mm"), legend.key.width = unit(7, "mm"),
        plot.tag = element_text(size = 14, face = "bold"))

f2_B <- ggplot(
  filter(genus_composition_data, site_group == "Forehead"),
  aes(Type, Abundance, fill = Genus)
) +
  geom_col(position = "fill", width = 0.88) +
  scale_fill_manual(values = genus_tax_plot$color) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), expand = c(0, 0)) +
  labs(x = NULL, y = "Forehead relative abundance (%)", fill = "Genus", tag = "B") +
  guides(fill = guide_legend(keyheight = unit(4.8, "mm"), keywidth = unit(7, "mm"))) +
  theme_manuscript() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        legend.text = element_text(size = 6),
        legend.key.height = unit(4.8, "mm"), legend.key.width = unit(7, "mm"),
        plot.tag = element_text(size = 14, face = "bold"))

f2_C <- ggplot(
  filter(selected_genus_data, site_group == "Antecubital"),
  aes(Type, Abundance * 100, fill = Type)
) +
  geom_boxplot(width = 0.68, outlier.shape = NA, colour = "black") +
  geom_point(aes(shape = subject_id), colour = "black",
             position = position_jitter(width = 0.08), size = 1.5) +
  facet_wrap(~Genus, nrow = 1, scales = "free_y") +
  scale_fill_manual(values = type_col, drop = FALSE) +
  labs(x = "Type", y = "Antecubital relative abundance (%)",
       fill = "Sampling type", shape = "Participant", tag = "C") +
  theme_manuscript() +
  theme(strip.text = element_text(face = "italic"), legend.position = "none",
        plot.tag = element_text(size = 14, face = "bold"))

f2_D <- ggplot(
  filter(selected_genus_data, site_group == "Forehead"),
  aes(Type, Abundance * 100, fill = Type)
) +
  geom_boxplot(width = 0.68, outlier.shape = NA, colour = "black") +
  geom_point(aes(shape = subject_id), colour = "black",
             position = position_jitter(width = 0.08), size = 1.5) +
  facet_wrap(~Genus, nrow = 1, scales = "free_y") +
  scale_fill_manual(values = type_col, drop = FALSE) +
  labs(x = "Type", y = "Forehead relative abundance (%)",
       fill = "Sampling type", shape = "Participant", tag = "D") +
  theme_manuscript() +
  theme(strip.text = element_text(face = "italic"), legend.position = "bottom",
        plot.tag = element_text(size = 14, face = "bold"))

# 3. Assemble and inspect
Figure2 <- (
  (f2_A | f2_B) /
    wrap_elements(full = f2_C) /
    wrap_elements(full = f2_D)
)

print(Figure2)
print(genus_composition_data)
print(selected_genus_data)

# 4. Save the inspected objects
ggsave(
  file.path(figure_dir, "Figure2_bacterial_community_composition.png"),
  plot = Figure2, width = 8, height = 8.8, dpi = 600,
  bg = "white", limitsize = FALSE
)
ggsave(
  file.path(figure_dir, "Figure2_bacterial_community_composition.tiff"),
  plot = Figure2, width = 8, height = 8.8, dpi = 600,
  compression = "lzw", bg = "white", limitsize = FALSE
)
write.csv(
  genus_composition_data,
  file.path(table_dir, "Figure2_genus_composition_data.csv"),
  row.names = FALSE
)
write.csv(
  selected_genus_data,
  file.path(table_dir, "Figure2_selected_genus_boxplot_data.csv"),
  row.names = FALSE
)

```

# Figure 3 — shared genera across sampling systems

```{r figure-3, message=FALSE, warning=FALSE}
# Generate two directly comparable three-panel figures:
# (1) no mean-abundance filtering and (2) mean abundance >=0.1%.

# 1. Data processing
upset_total_unfiltered <- prepare_upset_data(
  phy, site = NULL, mean_abundance_cutoff = 0
)
upset_antecubital_unfiltered <- prepare_upset_data(
  phy, site = "Antecubital", mean_abundance_cutoff = 0
)
upset_forehead_unfiltered <- prepare_upset_data(
  phy, site = "Forehead", mean_abundance_cutoff = 0
)

upset_total_filtered <- prepare_upset_data(
  phy, site = NULL, mean_abundance_cutoff = 0.001
)
upset_antecubital_filtered <- prepare_upset_data(
  phy, site = "Antecubital", mean_abundance_cutoff = 0.001
)
upset_forehead_filtered <- prepare_upset_data(
  phy, site = "Forehead", mean_abundance_cutoff = 0.001
)

Figure3_summary <- bind_rows(
  mutate(upset_total_unfiltered$summary, filtering = "No abundance filtering"),
  mutate(upset_antecubital_unfiltered$summary, filtering = "No abundance filtering"),
  mutate(upset_forehead_unfiltered$summary, filtering = "No abundance filtering"),
  mutate(upset_total_filtered$summary, filtering = "Mean abundance >=0.1%"),
  mutate(upset_antecubital_filtered$summary, filtering = "Mean abundance >=0.1%"),
  mutate(upset_forehead_filtered$summary, filtering = "Mean abundance >=0.1%")
) %>%
  mutate(
    subset = recode(
      subset,
      Antecubital = "Antecubital fossa"
    )
  )

Figure3_membership <- bind_rows(
  mutate(upset_total_unfiltered$data, subset = "Total", filtering = "No abundance filtering"),
  mutate(upset_antecubital_unfiltered$data, subset = "Antecubital fossa", filtering = "No abundance filtering"),
  mutate(upset_forehead_unfiltered$data, subset = "Forehead", filtering = "No abundance filtering"),
  mutate(upset_total_filtered$data, subset = "Total", filtering = "Mean abundance >=0.1%"),
  mutate(upset_antecubital_filtered$data, subset = "Antecubital fossa", filtering = "Mean abundance >=0.1%"),
  mutate(upset_forehead_filtered$data, subset = "Forehead", filtering = "Mean abundance >=0.1%")
)

# 2. Visualisation
Figure3_unfiltered <- (
  wrap_elements(full = upset_plot(upset_total_unfiltered, "Total", "genera")) /
    wrap_elements(full = upset_plot(upset_antecubital_unfiltered, "Antecubital fossa", "genera")) /
    wrap_elements(full = upset_plot(upset_forehead_unfiltered, "Forehead", "genera"))
) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 14, face = "bold"))

Figure3_filtered <- (
  wrap_elements(full = upset_plot(upset_total_filtered, "Total", "genera")) /
    wrap_elements(full = upset_plot(upset_antecubital_filtered, "Antecubital fossa", "genera")) /
    wrap_elements(full = upset_plot(upset_forehead_filtered, "Forehead", "genera"))
) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 14, face = "bold"))

# 3. Inspect
print(Figure3_unfiltered)
print(Figure3_filtered)
print(Figure3_summary)
print(Figure3_membership)

# 4. Save the inspected objects
png(
  file.path(figure_dir, "Figure3_UpSet_Genus_unfiltered.png"),
  width = 7.5, height = 11.5, units = "in", res = 600, bg = "white"
)
print(Figure3_unfiltered)
dev.off()

png(
  file.path(figure_dir, "Figure3_UpSet_Genus_0.1pct.png"),
  width = 7.5, height = 11.5, units = "in", res = 600, bg = "white"
)
print(Figure3_filtered)
dev.off()

tiff(
  file.path(figure_dir, "Figure3_UpSet_Genus_unfiltered.tiff"),
  width = 7.5, height = 11.5, units = "in", res = 600,
  compression = "lzw", bg = "white"
)
print(Figure3_unfiltered)
dev.off()

tiff(
  file.path(figure_dir, "Figure3_UpSet_Genus_0.1pct.tiff"),
  width = 7.5, height = 11.5, units = "in", res = 600,
  compression = "lzw", bg = "white"
)
print(Figure3_filtered)
dev.off()

write.csv(
  Figure3_summary,
  file.path(table_dir, "Figure3_Genus_shared_all_five_summary.csv"),
  row.names = FALSE
)
write.csv(
  Figure3_membership,
  file.path(table_dir, "Figure3_Genus_membership.csv"),
  row.names = FALSE
)

```

# Figure S1 — participant-level taxonomic composition

```{r figure-s1, message=FALSE, warning=FALSE}
# 5. Figure S1: participant-level composition at three taxonomic ranks
# -----------------------------------------------------------------------------

# Genus and Species use the existing taxa_plot() function directly so the
# phylum-specific palette and Other grouping exactly follow functions.R.
# 1. Data processing
composition_ranks <- c("Phylum", "Genus", "Species")
participant_composition <- list()
participant_colours <- list()
major_taxa_by_rank <- list()

# Phylum panel: one colour from the same phylum palettes used by taxa_plot().
phy_phylum <- tax_glom(phy, taxrank = "Phylum", NArm = FALSE)
phy_phylum_relative <- transform_sample_counts(phy_phylum, function(x) x / sum(x))
major_phylum_ids <- taxa_names(phy_phylum_relative)[
  taxa_sums(phy_phylum_relative) / nsamples(phy_phylum_relative) > 0.01
]
phylum_data <- psmelt(phy_phylum) %>%
  mutate(Phylum = ifelse(OTU %in% major_phylum_ids, as.character(Phylum), "Other")) %>%
  left_join(swab_key, by = "swab_type2")
phylum_palette_name <- c(
  Actinobacteriota = "Reds", Actinomycetota = "Reds",
  Firmicutes = "Blues", Bacillota = "Blues",
  Bacteroidota = "Purples", Proteobacteria = "Greens",
  Pseudomonadota = "Greens", Fusobacteriota = "YlOrBr"
)
major_phyla <- setdiff(sort(unique(phylum_data$Phylum)), "Other")
phylum_colours <- setNames(vapply(major_phyla, function(phylum) {
  palette_name <- if (phylum %in% names(phylum_palette_name)) {
    phylum_palette_name[[phylum]]
  } else {
    "BrBG"
  }
  rev(brewer.pal(9, palette_name))[5]
}, character(1)), major_phyla)
phylum_colours <- c(phylum_colours, Other = "#D3D3D3")
phylum_data$Phylum <- factor(phylum_data$Phylum, levels = names(phylum_colours))

# Genus and Species panels use the original taxa_plot() result directly.
phy_genus_s1 <- tax_glom(phy, taxrank = "Genus", NArm = FALSE)
phy_genus_s1_relative <- transform_sample_counts(phy_genus_s1, function(x) x / sum(x))
major_genus_s1_ids <- taxa_names(phy_genus_s1_relative)[
  taxa_sums(phy_genus_s1_relative) / nsamples(phy_genus_s1_relative) > 0.01
]
genus_s1 <- taxa_plot(
  psmelt(phy_genus_s1), "Genus", major_genus_s1_ids,
  "swab_type2", "Actinobacteriota"
)
genus_s1_data <- genus_s1$data$df %>% left_join(swab_key, by = "swab_type2")

phy_species_s1 <- tax_glom(phy, taxrank = "Species", NArm = FALSE)
phy_species_s1_relative <- transform_sample_counts(phy_species_s1, function(x) x / sum(x))
major_species_s1_ids <- taxa_names(phy_species_s1_relative)[
  taxa_sums(phy_species_s1_relative) / nsamples(phy_species_s1_relative) > 0.01
]
species_s1 <- taxa_plot(
  psmelt(phy_species_s1), "Species", major_species_s1_ids,
  "swab_type2", "Actinobacteriota"
)
species_s1_data <- species_s1$data$df %>% left_join(swab_key, by = "swab_type2")

participant_composition <- list(
  Phylum = phylum_data, Genus = genus_s1_data, Species = species_s1_data
)
participant_colours <- list(
  Phylum = phylum_colours, Genus = genus_s1$color, Species = species_s1$color
)
major_taxa_by_rank <- list(
  Phylum = unique(as.character(tax_table(phy_phylum)[major_phylum_ids, "Phylum"])),
  Genus = unique(as.character(tax_table(phy_genus_s1)[major_genus_s1_ids, "Genus"])),
  Species = unique(as.character(tax_table(phy_species_s1)[major_species_s1_ids, "Species"]))
)

# 2. Visualisation
s1_A <- ggplot(phylum_data, aes(Type, Abundance, fill = Phylum)) +
  geom_col(position = "fill", width = 0.92) +
  facet_grid(site_group ~ subject_id) +
  scale_fill_manual(values = phylum_colours, drop = FALSE) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), expand = c(0, 0)) +
  labs(x = NULL, y = "Relative abundance (%)", fill = "Phylum") +
  theme_manuscript(base_size = 7) +
  theme(axis.text.x = element_text(size = 6),
        strip.background = element_rect(fill = "white", colour = "black"),
        legend.text = element_text(size = 6))

s1_B <- ggplot(genus_s1_data, aes(Type, Abundance, fill = Genus)) +
  geom_col(position = "fill", width = 0.92) +
  facet_grid(site_group ~ subject_id) +
  scale_fill_manual(values = genus_s1$color, drop = FALSE) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), expand = c(0, 0)) +
  labs(x = NULL, y = "Relative abundance (%)", fill = "Genus") +
  theme_manuscript(base_size = 7) +
  theme(axis.text.x = element_text(size = 6),
        strip.background = element_rect(fill = "white", colour = "black"),
        legend.text = element_text(size = 6))

s1_C <- ggplot(species_s1_data, aes(Type, Abundance, fill = Species)) +
  geom_col(position = "fill", width = 0.92) +
  facet_grid(site_group ~ subject_id) +
  scale_fill_manual(values = species_s1$color, drop = FALSE) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), expand = c(0, 0)) +
  labs(x = NULL, y = "Relative abundance (%)", fill = "Species") +
  theme_manuscript(base_size = 7) +
  theme(axis.text.x = element_text(size = 6),
        strip.background = element_rect(fill = "white", colour = "black"),
        legend.text = element_text(size = 6))

FigureS1 <- (s1_A / s1_B / s1_C) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 16, face = "bold"))

# 3. Inspect
print(FigureS1)
print(major_taxa_by_rank)

# 4. Save the inspected object
ggsave(
  file.path(figure_dir, "FigureS1_participant_composition.png"),
  plot = FigureS1, width = 10.7, height = 10.8, dpi = 600,
  bg = "white", limitsize = FALSE
)
ggsave(
  file.path(figure_dir, "FigureS1_participant_composition.tiff"),
  plot = FigureS1, width = 10.7, height = 10.8, dpi = 600,
  compression = "lzw", bg = "white", limitsize = FALSE
)

```

# Figure S2 — UniFrac distance clustering

```{r tables-s3-s6, message=FALSE, warning=FALSE}
# Tables S3-S6 use the same taxa as Figure S1: global mean relative abundance >1%.
# "Others" is included in composition tables only, not in pairwise testing.
# 1. Data processing
type_levels <- paste0("Type ", 1:5)
type_columns <- paste0("Type", 1:5)
rank_relative_data <- list()

for (rank in composition_ranks) {
  ps_rank <- tax_glom(phy, taxrank = rank, NArm = FALSE)
  ps_rank <- transform_sample_counts(ps_rank, function(x) x / sum(x))
  rank_data <- psmelt(ps_rank) %>%
    left_join(swab_key, by = "swab_type2") %>%
    transmute(
      Sample,
      subject_id,
      site_group,
      Type,
      Rank = rank,
      Taxa = as.character(.data[[rank]]),
      Abundance = Abundance * 100
    )
  rank_relative_data[[rank]] <- rank_data
}
relative_long <- bind_rows(rank_relative_data)

# 2. Relative-abundance tables
relative_tables <- list()
pairwise_tables <- list()
comparisons <- combn(type_levels, 2, simplify = FALSE)

for (site in c("Antecubital", "Forehead")) {
  site_rows <- list()

  for (rank in composition_ranks) {
    selected_taxa <- major_taxa_by_rank[[rank]]
    current <- relative_long %>%
      filter(site_group == site, Rank == rank)

    selected <- current %>%
      filter(Taxa %in% selected_taxa) %>%
      group_by(Rank, Taxa, Type) %>%
      summarise(Relative_abundance = mean(Abundance), .groups = "drop")

    others <- current %>%
      group_by(Rank, Sample, Type) %>%
      summarise(
        Relative_abundance = 100 - sum(Abundance[Taxa %in% selected_taxa]),
        .groups = "drop"
      ) %>%
      group_by(Rank, Type) %>%
      summarise(Relative_abundance = mean(Relative_abundance), .groups = "drop") %>%
      mutate(Taxa = "Others")

    site_rows[[rank]] <- bind_rows(selected, others)
  }

  relative_tables[[site]] <- bind_rows(site_rows) %>%
    mutate(Type = gsub(" ", "", as.character(Type))) %>%
    pivot_wider(
      names_from = Type,
      values_from = Relative_abundance,
      values_fill = 0
    ) %>%
    mutate(Average = rowMeans(across(all_of(type_columns)))) %>%
    arrange(factor(Rank, levels = composition_ranks), desc(Average)) %>%
    mutate(across(where(is.numeric), ~round(.x, 2)))

  # 3. Paired taxon comparisons
  test_data <- relative_long %>%
    filter(
      site_group == site,
      Taxa %in% unlist(major_taxa_by_rank, use.names = FALSE)
    )

  test_rows <- list()
  row_index <- 1

  for (rank in composition_ranks) {
    rank_taxa <- major_taxa_by_rank[[rank]]

    for (taxon in rank_taxa) {
      current <- test_data %>%
        filter(Rank == rank, Taxa == taxon)

      result_row <- data.frame(Rank = rank, Taxa = taxon)

      for (comparison in comparisons) {
        paired_data <- current %>%
          filter(Type %in% comparison) %>%
          select(subject_id, Type, Abundance) %>%
          distinct() %>%
          pivot_wider(names_from = Type, values_from = Abundance)

        p_value <- NA_real_
        if (
          all(comparison %in% colnames(paired_data)) &&
          nrow(paired_data) > 1
        ) {
          differences <- paired_data[[comparison[1]]] -
            paired_data[[comparison[2]]]
          if (any(is.finite(differences)) && any(differences != 0, na.rm = TRUE)) {
            p_value <- tryCatch(
              wilcox.test(
                paired_data[[comparison[1]]],
                paired_data[[comparison[2]]],
                paired = TRUE,
                exact = FALSE
              )$p.value,
              error = function(e) NA_real_
            )
          }
        }

        column_name <- paste(comparison, collapse = " vs ")
        result_row[[column_name]] <- p_value
      }

      test_rows[[row_index]] <- result_row
      row_index <- row_index + 1
    }
  }

  pairwise_tables[[site]] <- bind_rows(test_rows) %>%
    mutate(
      across(
        -c(Rank, Taxa),
        ~p.adjust(.x, method = "BH")
      )
    ) %>%
    mutate(across(where(is.numeric), ~round(.x, 3)))
}

TableS3 <- relative_tables$Antecubital
TableS4 <- relative_tables$Forehead
TableS5 <- pairwise_tables$Antecubital
TableS6 <- pairwise_tables$Forehead

# 4. Inspect
print(TableS3)
print(TableS4)
print(TableS5)
print(TableS6)

# 5. Save the inspected tables
write.csv(
  TableS3,
  file.path(table_dir, "TableS3_Antecubital_relative_abundance_taxa_over_1pct.csv"),
  row.names = FALSE, na = "N/A"
)
write.csv(
  TableS4,
  file.path(table_dir, "TableS4_Forehead_relative_abundance_taxa_over_1pct.csv"),
  row.names = FALSE, na = "N/A"
)
write.csv(
  TableS5,
  file.path(table_dir, "TableS5_Antecubital_pairwise_taxa_over_1pct.csv"),
  row.names = FALSE, na = "N/A"
)
write.csv(
  TableS6,
  file.path(table_dir, "TableS6_Forehead_pairwise_taxa_over_1pct.csv"),
  row.names = FALSE, na = "N/A"
)

```

```{r figure-s2, message=FALSE, warning=FALSE}
# 6. Figure S2: distance heatmaps by site
# -----------------------------------------------------------------------------

subject_annotation_col <- subject_col
swab_annotation_col <- type_col

# 1. Data processing and heatmap construction
heatmap_settings <- data.frame(
  panel = c("A", "B", "C", "D"),
  site = c("Antecubital", "Antecubital", "Forehead", "Forehead"),
  method = c("unifrac", "wunifrac", "unifrac", "wunifrac"),
  title = c(
    "Antecubital Unweighted UniFrac",
    "Antecubital Weighted UniFrac",
    "Forehead Unweighted UniFrac",
    "Forehead Weighted UniFrac"
  )
)

heatmap_panels <- list()

for (i in seq_len(nrow(heatmap_settings))) {
  current <- heatmap_settings[i, ]
  keep <- sample_names(phy_rar)[
    as.character(sample_data(phy_rar)$site_group) == current$site
  ]
  ps_site <- prune_samples(keep, phy_rar)
  ps_site <- prune_taxa(taxa_sums(ps_site) > 0, ps_site)

  distance_matrix <- as.matrix(
    phyloseq::distance(ps_site, method = current$method)
  )
  metadata <- data.frame(sample_data(ps_site), check.names = FALSE) %>%
    rownames_to_column("sample_name") %>%
    left_join(swab_key, by = "swab_type2")

  annotation <- data.frame(
    Subject = metadata$subject_id,
    Swab_type = metadata$Type
  )
  rownames(annotation) <- metadata$sample_name

  heatmap_panels[[current$panel]] <- wrap_elements(
    full = pheatmap(
      distance_matrix,
      silent = TRUE,
      clustering_method = "complete",
      annotation_col = annotation,
      annotation_row = annotation,
      annotation_colors = list(
        Subject = subject_col,
        Swab_type = type_col
      ),
      show_rownames = TRUE,
      show_colnames = TRUE,
      fontsize = 6,
      border_color = NA,
      main = current$title
    )$gtable
  )
}

# 2. Assemble and inspect
FigureS2 <- (
  (heatmap_panels$A | heatmap_panels$B) /
    (heatmap_panels$C | heatmap_panels$D)
) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 16, face = "bold"))

print(FigureS2)

# 3. Save the inspected object
ggsave(
  file.path(figure_dir, "FigureS2_UniFrac_heatmaps.png"),
  plot = FigureS2, width = 10.5, height = 8.5, dpi = 600,
  bg = "white", limitsize = FALSE
)
ggsave(
  file.path(figure_dir, "FigureS2_UniFrac_heatmaps.tiff"),
  plot = FigureS2, width = 10.5, height = 8.5, dpi = 600,
  compression = "lzw", bg = "white", limitsize = FALSE
)

```

# Figure S3 — participant-level alpha and beta diversity

```{r figure-s3, message=FALSE, warning=FALSE}
# 7. Figure S3: participant-level alpha and beta diversity
# -----------------------------------------------------------------------------

# 1. Data processing
alpha_s3 <- estimate_richness(
  phy_rar, measures = c("Shannon", "Observed")
) %>%
  rownames_to_column("sample_name") %>%
  mutate(sample_name = sub("^X", "", sample_name)) %>%
  left_join(meta_rar, by = "sample_name") %>%
  pivot_longer(
    c(Shannon, Observed),
    names_to = "Index",
    values_to = "Value"
  ) %>%
  mutate(
    Index = factor(
      Index,
      levels = c("Shannon", "Observed"),
      labels = c("Shannon index", "Observed ASVs")
    )
  )

FigureS3_alpha_statistics <- ggpubr::compare_means(
  Value ~ subject_id,
  data = alpha_s3,
  group.by = c("site_group", "Index"),
  method = "wilcox.test",
  p.adjust.method = "BH"
)

# 2. Alpha-diversity plots
s3_A <- alpha_plot(
  filter(alpha_s3, site_group == "Antecubital"),
  x_var = "subject_id",
  fill_var = "subject_id",
  shape_var = "Type",
  site_title = "Antecubital fossa",
  y_title = "Alpha diversity",
  paired = FALSE,
  show_legend = TRUE
) + scale_fill_manual(values = subject_col, drop = FALSE)

s3_B <- alpha_plot(
  filter(alpha_s3, site_group == "Forehead"),
  x_var = "subject_id",
  fill_var = "subject_id",
  shape_var = "Type",
  site_title = "Forehead",
  y_title = "Alpha diversity",
  paired = FALSE,
  show_legend = FALSE
) + scale_fill_manual(values = subject_col, drop = FALSE)

# 3. Beta-diversity statistics and plots
s3_C_result <- beta_plot(
  phy_rar, "Antecubital", "unifrac", "subject_id",
  "subject_id", "Type", NULL,
  "Antecubital fossa", "Unweighted UniFrac", FALSE
)
s3_D_result <- beta_plot(
  phy_rar, "Antecubital", "wunifrac", "subject_id",
  "subject_id", "Type", NULL,
  "Antecubital fossa", "Weighted UniFrac", TRUE
)
s3_E_result <- beta_plot(
  phy_rar, "Forehead", "unifrac", "subject_id",
  "subject_id", "Type", NULL,
  "Forehead", "Unweighted UniFrac", FALSE
)
s3_F_result <- beta_plot(
  phy_rar, "Forehead", "wunifrac", "subject_id",
  "subject_id", "Type", NULL,
  "Forehead", "Weighted UniFrac", TRUE
)

s3_C <- s3_C_result$plot +
  scale_colour_manual(values = subject_col, drop = FALSE)
s3_D <- s3_D_result$plot +
  scale_colour_manual(values = subject_col, drop = FALSE)
s3_E <- s3_E_result$plot +
  scale_colour_manual(values = subject_col, drop = FALSE)
s3_F <- s3_F_result$plot +
  scale_colour_manual(values = subject_col, drop = FALSE)

FigureS3_PERMANOVA <- bind_rows(
  s3_C_result$permanova,
  s3_D_result$permanova,
  s3_E_result$permanova,
  s3_F_result$permanova
)

# 4. Assemble and inspect
FigureS3 <- (
  wrap_elements(full = (s3_A | s3_B)) /
    wrap_elements(full = (s3_C | s3_D)) /
    wrap_elements(full = (s3_E | s3_F))
) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 16, face = "bold"))

print(FigureS3)
print(FigureS3_alpha_statistics)
print(FigureS3_PERMANOVA)

# 5. Save the inspected objects
ggsave(
  file.path(figure_dir, "FigureS3_participant_alpha_beta.png"),
  plot = FigureS3, width = 10, height = 10, dpi = 600,
  bg = "white", limitsize = FALSE
)
ggsave(
  file.path(figure_dir, "FigureS3_participant_alpha_beta.tiff"),
  plot = FigureS3, width = 10, height = 10, dpi = 600,
  compression = "lzw", bg = "white", limitsize = FALSE
)
write.csv(
  FigureS3_alpha_statistics,
  file.path(table_dir, "FigureS3_alpha_pairwise_Wilcoxon_BH.csv"),
  row.names = FALSE
)
write.csv(
  FigureS3_PERMANOVA,
  file.path(table_dir, "FigureS3_PERMANOVA.csv"),
  row.names = FALSE
)
write.csv(
  swab_key,
  file.path(table_dir, "Swab_type_key.csv"),
  row.names = FALSE
)

```

# Final figure preview

```{r final-figure-preview, echo=FALSE, message=FALSE, warning=FALSE}
final_png <- c(
  "Figure1_bacterial_diversity.png",
  "Figure2_bacterial_community_composition.png",
  "Figure3_UpSet_Genus_0.1pct.png",
  "FigureS1_participant_composition.png",
  "FigureS2_UniFrac_heatmaps.png",
  "FigureS3_participant_alpha_beta.png"
)
knitr::include_graphics(file.path(figure_dir, final_png))
```

# Table 3. PERMANOVA results for bacterial community variation based on weighted UniFrac distance (9,999 permutations). 

```{r}
# Fri Aug 14 10:51:08 2026 ------------------------------


set.seed(42)
dist <- phyloseq::distance(phy_rar, method = "wunifrac")
dist2 <- phyloseq::distance(phy_rar, method = "unifrac")
meta   <- phy_rar %>% sample_data() %>% data.frame()

meta$swab_type

meta2 <- meta %>% 
  mutate(
    `Collection_medium` = case_when(
      swab_type2 %in% c("Puritan_gel", "Copan") ~  "Gel",
      swab_type2 %in% "Puritan_buffer" ~ "buffer",
      swab_type2 %in% "Omni" ~ "Omni_buffer",
      swab_type2 %in% "Eswab" ~ "Eswab_buffer"
      ),
    `Swab_material` = case_when(
      swab_type %in%  "Eswab" ~  "Nylon",
      swab_type %in%  "Omni" ~ "Polyester",
      swab_type %in% "Puritan" ~ "Polyurethane foam",
      swab_type %in% "Copan" ~"Rayon"
      )
    )      

```


```{r}
# # PERMANOVA
# set.seed(42)
# Perm1    <- adonis2(
#   dist ~ subject_id + swab_type2 + site_group + Collection_medium + Swab_material,
#   data         = meta2,
#   permutations = 9999,
#   # strata = meta2$SubjectID,
#   by ="margin"
# ) 
# 
# set.seed(42)
# 
# Perm <- adonis2(
#   dist ~ subject_id + swab_type2 + site_group +  Collection_medium + Swab_material,
#   data = meta2,
#   permutations = 9999,
#   strata = meta2$SubjectID,
#   by = "terms"
# )
# 
# 
# set.seed(42)
# Perm2    <- adonis2(
#   dist ~ subject_id + swab_type2 + site_group + Collection_medium + Swab_material,
#   data         = meta2,
#   permutations = 9999,
#   # strata = meta2$SubjectID,
#   # by ="margin"
#     by = "terms"
# ) 
#  
```


```{r}




# collection medium만
df5 <-adonis2(
  dist ~ Collection_medium,
  data = meta2,
  permutations = 9999,
  strata = meta2$subject_id
)


# swab material만
df1 <-adonis2(
  dist ~ subject_id,
  data = meta2,
  permutations = 9999)


# site만
df2 <-adonis2(
  dist ~ site_group,
  data = meta2,
  permutations = 9999,
  strata = meta2$subject_id
)
# swab type만
df3 <- adonis2(
  dist ~ swab_type2,
  data = meta2,
  permutations = 9999,
  strata = meta2$subject_id
)
# swab material만
df4 <-adonis2(
  dist ~ Swab_material,
  data = meta2,
  permutations = 9999,
  strata = meta2$subject_id
)

univar_per <- data.frame(
  variable = c("Participant identity", "Anatomical site", "Sampling system", "Swab-tip material", "Collection medium"),
  df = c(df1$Df[1], df2$Df[1],  df3$Df[1], df4$Df[1], df5$Df[1]),
  `F` = c(df1$`F`[1], df2$`F`[1],  df3$`F`[1], df4$`F`[1], df5$`F`[1]),
  R2 = c(df1$R2[1], df2$R2[1], df3$R2[1], df4$R2[1], df5$R2[1]),
  `p-value` = c(df1$`Pr(>F)`[1], df2$`Pr(>F)`[1], df3$`Pr(>F)`[1], df4$`Pr(>F)`[1], df5$`Pr(>F)`[1])
)


univar_per$q.value <- p.adjust(univar_per$p.value, "fdr")

write.csv(
    univar_per,
    file.path(table_dir, "Table3_permanova_BH.csv"),
    row.names = FALSE
  )
```
# Depth 
```{r}

sum(sample_sums(phy)) # 847074

summary(sample_sums(phy))

sd(sample_sums(phy)) 
  # Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  #  5876   12330   17413   16941   21164   31707 
# Table S3: sequencing depth and observed taxonomic richness
# Species-level output is intentionally excluded from the revised manuscript.

# 1. Overall summary
ps_sub <- prune_taxa(taxa_sums(phy) > 0, phy)
reads <- sample_sums(ps_sub)
tax_df <- as.data.frame(tax_table(ps_sub), stringsAsFactors = FALSE)

depth_rows <- list(
  data.frame(
    Scope = "Overall",
    Site = "All",
    Sampling_system = "All",
    N_samples = nsamples(ps_sub),
    Total_reads = sum(reads),
    Mean_Reads = round(mean(reads), 1),
    SD_Reads = round(sd(reads), 1),
    ASV_Count = ntaxa(ps_sub),
    Phylum_Count = n_distinct(na.omit(tax_df$Phylum)),
    Genus_Count = n_distinct(na.omit(tax_df$Genus))
  )
)

# 2. Anatomical-site summaries
row_index <- 2
for (site_value in c("Ac", "Fh")) {
  keep <- sample_names(phy)[
    as.character(sample_data(phy)$site_id) == site_value
  ]
  ps_sub <- prune_samples(keep, phy)
  ps_sub <- prune_taxa(taxa_sums(ps_sub) > 0, ps_sub)
  reads <- sample_sums(ps_sub)
  tax_df <- as.data.frame(tax_table(ps_sub), stringsAsFactors = FALSE)

  depth_rows[[row_index]] <- data.frame(
    Scope = "Anatomical site",
    Site = site_value,
    Sampling_system = "All",
    N_samples = nsamples(ps_sub),
    Total_reads = sum(reads),
    Mean_Reads = round(mean(reads), 1),
    SD_Reads = round(sd(reads), 1),
    ASV_Count = ntaxa(ps_sub),
    Phylum_Count = n_distinct(na.omit(tax_df$Phylum)),
    Genus_Count = n_distinct(na.omit(tax_df$Genus))
  )
  row_index <- row_index + 1
}

# 3. Site-by-sampling-system summaries
for (site_value in c("Ac", "Fh")) {
  for (i in seq_len(nrow(swab_key))) {
    swab_value <- swab_key$swab_type2[i]
    type_value <- as.character(swab_key$Type[i])
    keep <- sample_names(phy)[
      as.character(sample_data(phy)$site_id) == site_value &
        as.character(sample_data(phy)$swab_type2) == swab_value
    ]
    ps_sub <- prune_samples(keep, phy)
    ps_sub <- prune_taxa(taxa_sums(ps_sub) > 0, ps_sub)
    reads <- sample_sums(ps_sub)
    tax_df <- as.data.frame(tax_table(ps_sub), stringsAsFactors = FALSE)

    depth_rows[[row_index]] <- data.frame(
      Scope = "Anatomical site x sampling system",
      Site = site_value,
      Sampling_system = type_value,
      N_samples = nsamples(ps_sub),
      Total_reads = sum(reads),
      Mean_Reads = round(mean(reads), 1),
      SD_Reads = round(sd(reads), 1),
      ASV_Count = ntaxa(ps_sub),
      Phylum_Count = n_distinct(na.omit(tax_df$Phylum)),
      Genus_Count = n_distinct(na.omit(tax_df$Genus))
    )
    row_index <- row_index + 1
  }
}

table_s3_depth <- bind_rows(depth_rows) %>%
  mutate(
    Site = factor(Site, levels = c("All", "Ac", "Fh")),
    Sampling_system = factor(
      Sampling_system,
      levels = c("All", paste0("Type ", 1:5))
    )
  ) %>%
  arrange(Site, Sampling_system) %>%
  mutate(
    Site = as.character(Site),
    Sampling_system = as.character(Sampling_system)
  )

# 4. Inspect
print(table_s3_depth)

# 5. Save the inspected table
write.csv(
  table_s3_depth,
  file.path(table_dir, "TableS3_sequencing_depth_by_site_and_sampling_system.csv"),
  row.names = FALSE
)

```

