---
title: "Filtering sensitivity analysis for the DT skin-swab study"
author: "DT Swab Study"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

# Analysis objective

This analysis evaluates how additional rare-feature filtering changes data retention,
alpha diversity, beta diversity, sampling-system effects, and feature sharing after
the prevalence-based decontamination step used in the main analysis. The saved
post-decontamination phyloseq object is the common starting point for every filter.

```{r packages}
suppressPackageStartupMessages({
  library(phyloseq)
  library(vegan)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(patchwork)
  library(ComplexUpset)
  library(pheatmap)
  library(RColorBrewer)
  library(ggpubr)
})

set.seed(42)

out_dir <- file.path(getwd(), "Filtering_sensitivity_output")
fig_dir <- file.path(out_dir, "Figures")
tab_dir <- file.path(out_dir, "Tables")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(tab_dir, recursive = TRUE, showWarnings = FALSE)

filter_levels <- c(
  "No additional filter",
  "Remove count <=1",
  "Remove count <=2",
  "Remove count <=10",
  "Prevalence >=2 samples",
  "Prevalence >=5 samples",
  "Prevalence >=10% samples",
  "Mean abundance >=0.1%",
  "Mean abundance >=1%"
)

filter_ids <- c(
  "no_filter",
  "remove_count_le_1",
  "remove_count_le_2",
  "remove_count_le_10",
  "prevalence_ge_2",
  "prevalence_ge_5",
  "prevalence_ge_10pct",
  "mean_abundance_ge_0.1pct",
  "mean_abundance_ge_1pct"
)

filter_labels <- setNames(filter_levels, filter_ids)

swab_key <- data.frame(
  swab_type2 = c("Copan", "Puritan_gel", "Puritan_buffer", "Omni", "Eswab"),
  Type = factor(paste0("Type ", 1:5), levels = paste0("Type ", 1:5))
)

type_col <- setNames(brewer.pal(5, "Set2"), paste0("Type ", 1:5))
metric_col <- c(
  "ASV" = "#377eb8",
  "Genus" = "#e41a1c",
  "Reads" = "#4daf4a"
)

theme_ms <- theme_bw(base_size = 9) +
  theme(
    axis.title = element_text(face = "bold", colour = "black"),
    axis.text = element_text(colour = "black"),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white", colour = "black"),
    strip.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    plot.margin = margin(6, 6, 6, 6)
  )

save_plot <- function(plot, stem, width, height) {
  ggsave(
    file.path(fig_dir, paste0(stem, ".png")), plot,
    width = width, height = height, dpi = 600,
    bg = "white", limitsize = FALSE
  )
  ggsave(
    file.path(fig_dir, paste0(stem, ".tiff")), plot,
    width = width, height = height, dpi = 600,
    compression = "lzw", bg = "white", limitsize = FALSE
  )
}
```

# Input and filter definitions

```{r input-and-filters}
phy <- readRDS("./Phyloseq/phy_F270R240_260812_v4.rds")
phy <- subset_samples(phy, control_status == "sample") %>%
  prune_taxa(taxa_sums(.) > 0, .)

target_depth <- 5876L

otu_matrix <- function(ps) {
  otu <- as(otu_table(ps), "matrix")
  if (!taxa_are_rows(ps)) otu <- t(otu)
  otu
}

augment_metadata <- function(ps) {
  md <- data.frame(sample_data(ps), check.names = FALSE) %>%
    rownames_to_column("sample_name") %>%
    left_join(swab_key, by = "swab_type2") %>%
    mutate(
      collection_medium = case_when(
        swab_type2 %in% c("Puritan_gel", "Copan") ~ "Gel",
        swab_type2 == "Puritan_buffer" ~ "Puritan buffer",
        swab_type2 == "Omni" ~ "Omni buffer",
        swab_type2 == "Eswab" ~ "ESwab buffer",
        TRUE ~ NA_character_
      ),
      swab_material = case_when(
        swab_type == "Eswab" ~ "Nylon",
        swab_type == "Omni" ~ "Polyester",
        swab_type == "Puritan" ~ "Polyurethane foam",
        swab_type == "Copan" ~ "Rayon",
        TRUE ~ NA_character_
      )
    )
  rownames(md) <- md$sample_name
  md
}

apply_feature_filter <- function(ps, filter_id) {
  otu <- otu_matrix(ps)
  total_count <- rowSums(otu)
  prevalence <- rowSums(otu > 0)
  rel <- sweep(otu, 2, colSums(otu), "/")
  rel[!is.finite(rel)] <- 0
  mean_rel_pct <- rowMeans(rel) * 100

  keep <- switch(
    filter_id,
    no_filter = total_count > 0,
    remove_count_le_1 = total_count > 1,
    remove_count_le_2 = total_count > 2,
    remove_count_le_10 = total_count > 10,
    prevalence_ge_2 = prevalence >= 2,
    prevalence_ge_5 = prevalence >= 5,
    prevalence_ge_10pct = prevalence >= ceiling(0.10 * nsamples(ps)),
    `mean_abundance_ge_0.1pct` = mean_rel_pct >= 0.1,
    `mean_abundance_ge_1pct` = mean_rel_pct >= 1,
    stop("Unknown filter: ", filter_id)
  )

  prune_taxa(keep, ps) %>% prune_taxa(taxa_sums(.) > 0, .)
}

filtered_ps <- setNames(
  lapply(filter_ids, function(id) apply_feature_filter(phy, id)),
  filter_ids
)

rarefy_fixed_depth <- function(ps, depth = target_depth) {
  keep_samples <- sample_names(ps)[sample_sums(ps) >= depth]
  if (length(keep_samples) < 2) return(NULL)
  ps_keep <- prune_samples(keep_samples, ps) %>%
    prune_taxa(taxa_sums(.) > 0, .)
  set.seed(42)
  rarefy_even_depth(
    ps_keep,
    sample.size = depth,
    rngseed = 42,
    replace = FALSE,
    trimOTUs = TRUE,
    verbose = FALSE
  )
}

filtered_rar <- lapply(filtered_ps, rarefy_fixed_depth)
```

# Data retention

```{r retention-summary}
base_reads <- sum(sample_sums(phy))
base_asvs <- ntaxa(phy)

retention_summary <- bind_rows(lapply(filter_ids, function(id) {
  ps <- filtered_ps[[id]]
  depths <- sample_sums(ps)
  ps_rar <- filtered_rar[[id]]
  tibble(
    filter_id = id,
    filter = filter_labels[[id]],
    retained_asvs = ntaxa(ps),
    removed_asvs = base_asvs - ntaxa(ps),
    asv_retained_pct = 100 * ntaxa(ps) / base_asvs,
    retained_reads = sum(depths),
    removed_reads = base_reads - sum(depths),
    read_retained_pct = 100 * sum(depths) / base_reads,
    samples_total = nsamples(ps),
    samples_below_5876 = sum(depths < target_depth),
    samples_rarefied_5876 = if (is.null(ps_rar)) 0L else nsamples(ps_rar),
    depth_min = min(depths),
    depth_median = median(depths),
    depth_max = max(depths)
  )
})) %>%
  mutate(filter = factor(filter, levels = filter_levels))

write.csv(
  retention_summary %>% mutate(filter = as.character(filter)),
  file.path(tab_dir, "filtering_retention_summary.csv"),
  row.names = FALSE
)

retention_long <- retention_summary %>%
  select(filter, asv_retained_pct, read_retained_pct) %>%
  pivot_longer(-filter, names_to = "metric", values_to = "retained_pct") %>%
  mutate(
    metric = recode(
      metric,
      asv_retained_pct = "ASV",
      read_retained_pct = "Reads"
    )
  )

p_retention <- ggplot(
  retention_long,
  aes(filter, retained_pct, colour = metric, group = metric)
) +
  geom_hline(yintercept = 100, colour = "grey80", linewidth = 0.4) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  scale_colour_manual(values = metric_col[c("ASV", "Reads")]) +
  scale_y_continuous(limits = c(0, 101)) +
  labs(
    x = NULL,
    y = "Retained (%)",
    colour = "Metric",
    title = "Data retained after additional feature filtering"
  ) +
  theme_ms +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_plot(p_retention, "Figure1_Filtering_retention", 8.5, 4.8)
retention_summary
```

# Alpha-diversity sensitivity

```{r alpha-diversity}
alpha_values <- bind_rows(lapply(filter_ids, function(id) {
  ps <- filtered_rar[[id]]
  if (is.null(ps)) return(tibble())
  estimate_richness(ps, measures = c("Shannon", "Observed")) %>%
    rownames_to_column("sample_name") %>%
    mutate(sample_name = sub("^X", "", sample_name)) %>%
    left_join(
      augment_metadata(ps) %>%
        select(sample_name, subject_id, site_group, swab_type2, Type),
      by = "sample_name"
    ) %>%
    mutate(filter_id = id, filter = filter_labels[[id]])
})) %>%
  mutate(filter = factor(filter, levels = filter_levels))

pairwise_wilcox_tidy <- function(df, metric_name) {
  empty_result <- tibble(
    group1 = character(),
    group2 = character(),
    p_adjusted = numeric()
  )
  if (nrow(df) == 0 || n_distinct(df$Type[!is.na(df$Type)]) < 2) {
    return(empty_result)
  }
  pw <- pairwise.wilcox.test(
    x = df[[metric_name]],
    g = df$Type,
    p.adjust.method = "bonferroni",
    exact = FALSE
  )$p.value
  if (is.null(pw)) return(empty_result)
  as.data.frame(as.table(pw), stringsAsFactors = FALSE) %>%
    filter(!is.na(Freq)) %>%
    transmute(group1 = as.character(Var1), group2 = as.character(Var2), p_adjusted = Freq)
}

alpha_pairwise <- bind_rows(lapply(filter_ids, function(id) {
  bind_rows(lapply(c("Antecubital", "Forehead"), function(site) {
    dat <- alpha_values %>%
      filter(filter_id == id, site_group == site)
    bind_rows(lapply(c("Shannon", "Observed"), function(metric_name) {
      pairwise_wilcox_tidy(dat, metric_name) %>%
        mutate(
          filter_id = id,
          filter = filter_labels[[id]],
          site_group = site,
          metric = metric_name,
          significant_0.05 = p_adjusted < 0.05
        )
    }))
  }))
}))

alpha_significance_summary <- alpha_pairwise %>%
  group_by(filter_id, filter, site_group, metric) %>%
  summarise(
    pairwise_tests = n(),
    significant_pairs = sum(significant_0.05, na.rm = TRUE),
    minimum_adjusted_p = min(p_adjusted, na.rm = TRUE),
    .groups = "drop"
  )

write.csv(alpha_values, file.path(tab_dir, "alpha_diversity_values.csv"), row.names = FALSE)
write.csv(alpha_pairwise, file.path(tab_dir, "alpha_pairwise_wilcoxon.csv"), row.names = FALSE)
write.csv(
  alpha_significance_summary,
  file.path(tab_dir, "alpha_pairwise_significance_summary.csv"),
  row.names = FALSE
)

alpha_significance_summary
```

# PERMANOVA sensitivity

Participant identity is tested without participant-restricted permutations. Tests
for anatomical site, sampling system, swab-tip material, and collection medium use
participant identity as the permutation stratum, matching the repeated-measures
structure used in the revised analysis.

```{r permanova}
distance_labels <- c(
  wunifrac = "Weighted UniFrac",
  unifrac = "Unweighted UniFrac",
  bray = "Bray-Curtis"
)

variable_labels <- c(
  subject_id = "Participant identity",
  site_group = "Anatomical site",
  swab_type2 = "Sampling system",
  swab_material = "Swab-tip material",
  collection_medium = "Collection medium"
)

run_adonis_term <- function(dist_obj, md, variable, seed_value = 42L) {
  formula_obj <- reformulate(variable, response = "dist_obj")
  set.seed(seed_value)
  fit <- tryCatch(
    {
      if (variable == "subject_id") {
        adonis2(formula_obj, data = md, permutations = 9999)
      } else {
        adonis2(
          formula_obj,
          data = md,
          permutations = 9999,
          strata = md$subject_id
        )
      }
    },
    error = function(e) e
  )
  if (inherits(fit, "error")) {
    return(tibble(
      Df = NA_real_, SumOfSqs = NA_real_, R2 = NA_real_,
      F = NA_real_, p_value = NA_real_, error = conditionMessage(fit)
    ))
  }
  tibble(
    Df = fit$Df[1],
    SumOfSqs = fit$SumOfSqs[1],
    R2 = fit$R2[1],
    F = fit$F[1],
    p_value = fit$`Pr(>F)`[1],
    error = NA_character_
  )
}

permanova_results <- bind_rows(lapply(seq_along(filter_ids), function(i) {
  id <- filter_ids[[i]]
  ps <- filtered_rar[[id]]
  if (is.null(ps)) return(tibble())
  md <- augment_metadata(ps)

  bind_rows(lapply(seq_along(distance_labels), function(j) {
    distance_id <- names(distance_labels)[[j]]
    dist_obj <- phyloseq::distance(ps, method = distance_id)
    bind_rows(lapply(seq_along(variable_labels), function(k) {
      variable <- names(variable_labels)[[k]]
      run_adonis_term(
        dist_obj, md, variable,
        seed_value = 42L + j * 10L + k
      ) %>%
        mutate(
          filter_id = id,
          filter = filter_labels[[id]],
          distance_id = distance_id,
          distance = distance_labels[[distance_id]],
          variable_id = variable,
          variable = variable_labels[[variable]],
          n_samples = nrow(md),
          permutation_strata = ifelse(variable_id == "subject_id", "None", "Participant identity")
        )
    }))
  }))
})) %>%
  mutate(
    filter = factor(filter, levels = filter_levels),
    variable = factor(variable, levels = unname(variable_labels)),
    distance = factor(distance, levels = unname(distance_labels))
  ) %>%
  select(
    filter_id, filter, distance_id, distance, variable_id, variable,
    n_samples, permutation_strata, Df, SumOfSqs, R2, F, p_value, error
  )

write.csv(
  permanova_results %>% mutate(across(c(filter, distance, variable), as.character)),
  file.path(tab_dir, "permanova_sensitivity.csv"),
  row.names = FALSE
)

write.csv(
  permanova_results %>%
    filter(variable_id == "swab_type2") %>%
    mutate(across(c(filter, distance, variable), as.character)),
  file.path(tab_dir, "permanova_sampling_system_summary.csv"),
  row.names = FALSE
)

p_permanova <- ggplot(
  permanova_results,
  aes(filter, R2 * 100, colour = variable, group = variable)
) +
  geom_line(linewidth = 0.55) +
  geom_point(aes(shape = p_value < 0.05), size = 1.8) +
  facet_wrap(~distance, scales = "free_y", ncol = 1) +
  scale_shape_manual(values = c(`TRUE` = 19, `FALSE` = 1), name = "p < 0.05") +
  labs(
    x = NULL,
    y = expression(PERMANOVA~R^2~"(%)"),
    colour = "Variable",
    title = "PERMANOVA effect sizes across feature filters"
  ) +
  theme_ms +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_plot(p_permanova, "Figure2_PERMANOVA_sensitivity", 9.2, 8.5)
permanova_results
```

# Pairwise beta-diversity partitioning

```{r pairwise-beta}
distance_to_pairs <- function(dist_obj, md, filter_key, distance_key) {
  mat <- as.matrix(dist_obj)
  idx <- which(upper.tri(mat), arr.ind = TRUE)
  out <- tibble(
    sample_1 = rownames(mat)[idx[, 1]],
    sample_2 = colnames(mat)[idx[, 2]],
    distance_value = mat[idx]
  ) %>%
    left_join(
      md %>%
        select(sample_name, subject_id, site_group, Type) %>%
        rename(
          sample_1 = sample_name,
          subject_1 = subject_id,
          site_1 = site_group,
          type_1 = Type
        ),
      by = "sample_1"
    ) %>%
    left_join(
      md %>%
        select(sample_name, subject_id, site_group, Type) %>%
        rename(
          sample_2 = sample_name,
          subject_2 = subject_id,
          site_2 = site_group,
          type_2 = Type
        ),
      by = "sample_2"
    ) %>%
    mutate(
      same_participant = subject_1 == subject_2,
      same_site = site_1 == site_2,
      same_system = type_1 == type_2,
      comparison_group = case_when(
        same_participant & same_site & !same_system ~ "Within participant/site, between systems",
        same_participant & !same_site ~ "Within participant, between sites",
        !same_participant & same_site ~ "Between participants, same site",
        !same_participant & !same_site ~ "Between participants and sites",
        TRUE ~ "Other within participant/site"
      ),
      filter_id = filter_key,
      filter = filter_labels[[filter_key]],
      distance_id = distance_key,
      distance = distance_labels[[distance_key]]
    )
  out
}

pairwise_beta <- bind_rows(lapply(filter_ids, function(id) {
  ps <- filtered_rar[[id]]
  if (is.null(ps)) return(tibble())
  md <- augment_metadata(ps)
  bind_rows(lapply(names(distance_labels), function(distance_id) {
    distance_to_pairs(
      phyloseq::distance(ps, method = distance_id),
      md,
      id,
      distance_id
    )
  }))
})) %>%
  mutate(
    filter = factor(filter, levels = filter_levels),
    distance = factor(distance, levels = unname(distance_labels))
  )

pairwise_beta_summary <- pairwise_beta %>%
  group_by(filter_id, filter, distance_id, distance, comparison_group) %>%
  summarise(
    n_pairs = n(),
    mean_distance = mean(distance_value),
    median_distance = median(distance_value),
    sd_distance = sd(distance_value),
    q25 = quantile(distance_value, 0.25),
    q75 = quantile(distance_value, 0.75),
    .groups = "drop"
  )

method_vs_participant <- pairwise_beta %>%
  filter(
    (same_participant & same_site & !same_system) |
      (!same_participant & same_site & same_system)
  ) %>%
  mutate(
    method_contrast = ifelse(
      same_participant,
      "Different systems within participant/site",
      "Same system between participants/site"
    )
  ) %>%
  group_by(filter_id, filter, distance_id, distance, method_contrast) %>%
  summarise(
    n_pairs = n(),
    mean_distance = mean(distance_value),
    median_distance = median(distance_value),
    sd_distance = sd(distance_value),
    .groups = "drop"
  )

write.csv(pairwise_beta, file.path(tab_dir, "beta_pairwise_distances.csv"), row.names = FALSE)
write.csv(
  pairwise_beta_summary,
  file.path(tab_dir, "beta_distance_partition_summary.csv"),
  row.names = FALSE
)
write.csv(
  method_vs_participant,
  file.path(tab_dir, "beta_method_vs_participant_summary.csv"),
  row.names = FALSE
)

p_beta_partition <- pairwise_beta %>%
  filter(comparison_group != "Other within participant/site") %>%
  ggplot(aes(filter, distance_value, fill = comparison_group)) +
  geom_boxplot(outlier.shape = NA, linewidth = 0.3) +
  facet_wrap(~distance, scales = "free_y", ncol = 1) +
  labs(
    x = NULL,
    y = "Pairwise distance",
    fill = "Pair comparison",
    title = "Participant-, site-, and sampling-system-related beta diversity"
  ) +
  theme_ms +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_plot(p_beta_partition, "Figure3_Beta_distance_partition", 11, 9)
pairwise_beta_summary
```

# Concordance of beta-diversity results across filters

```{r beta-concordance}
distance_vectors <- lapply(filter_ids, function(id) {
  ps <- filtered_rar[[id]]
  if (is.null(ps)) return(NULL)
  setNames(
    lapply(names(distance_labels), function(distance_id) {
      as.vector(phyloseq::distance(ps, method = distance_id))
    }),
    names(distance_labels)
  )
})
names(distance_vectors) <- filter_ids

beta_concordance <- bind_rows(lapply(names(distance_labels), function(distance_id) {
  base_vec <- distance_vectors[["no_filter"]][[distance_id]]
  bind_rows(lapply(filter_ids, function(id) {
    candidate <- distance_vectors[[id]][[distance_id]]
    if (length(base_vec) != length(candidate)) {
      return(tibble(
        filter_id = id,
        distance_id = distance_id,
        spearman_rho = NA_real_,
        mantel_r = NA_real_,
        mantel_p = NA_real_,
        note = "Different sample sets after rarefaction"
      ))
    }
    ps <- filtered_rar[[id]]
    base_ps <- filtered_rar[["no_filter"]]
    base_dist <- phyloseq::distance(base_ps, method = distance_id)
    candidate_dist <- phyloseq::distance(ps, method = distance_id)
    set.seed(42)
    mt <- vegan::mantel(base_dist, candidate_dist, method = "spearman", permutations = 9999)
    tibble(
      filter_id = id,
      distance_id = distance_id,
      spearman_rho = suppressWarnings(cor(base_vec, candidate, method = "spearman")),
      mantel_r = unname(mt$statistic),
      mantel_p = mt$signif,
      note = NA_character_
    )
  }))
})) %>%
  mutate(
    filter = factor(filter_labels[filter_id], levels = filter_levels),
    distance = factor(distance_labels[distance_id], levels = unname(distance_labels))
  )

write.csv(
  beta_concordance %>% mutate(across(c(filter, distance), as.character)),
  file.path(tab_dir, "beta_distance_filter_concordance.csv"),
  row.names = FALSE
)

p_concordance <- ggplot(
  beta_concordance,
  aes(filter, distance, fill = spearman_rho)
) +
  geom_tile(colour = "white") +
  geom_text(aes(label = ifelse(is.na(spearman_rho), "NA", sprintf("%.3f", spearman_rho))), size = 2.7) +
  scale_fill_gradient2(
    low = "#2166ac", mid = "white", high = "#b2182b",
    midpoint = 0.5, limits = c(0, 1), na.value = "grey80"
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Spearman rho",
    title = "Concordance of beta-diversity distances with the unfiltered dataset"
  ) +
  theme_ms +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_plot(p_concordance, "Figure4_Beta_distance_concordance", 9.2, 4.5)
beta_concordance
```

# Community-composition stability

```{r composition-stability}
sample_composition_similarity <- function(base_ps, filtered_ps, filter_id) {
  base_otu <- otu_matrix(base_ps)
  filtered_otu <- otu_matrix(filtered_ps)
  all_taxa <- rownames(base_otu)
  aligned <- matrix(
    0,
    nrow = length(all_taxa),
    ncol = ncol(filtered_otu),
    dimnames = list(all_taxa, colnames(filtered_otu))
  )
  aligned[rownames(filtered_otu), ] <- filtered_otu
  common_samples <- intersect(colnames(base_otu), colnames(aligned))
  bind_rows(lapply(common_samples, function(sample_id) {
    x <- base_otu[, sample_id]
    y <- aligned[, sample_id]
    x <- x / sum(x)
    y <- y / sum(y)
    bc <- as.numeric(vegdist(rbind(x, y), method = "bray"))
    tibble(
      filter_id = filter_id,
      filter = filter_labels[[filter_id]],
      sample_name = sample_id,
      bray_curtis_to_unfiltered = bc,
      composition_similarity = 1 - bc
    )
  }))
}

composition_stability <- bind_rows(lapply(filter_ids, function(id) {
  sample_composition_similarity(phy, filtered_ps[[id]], id)
})) %>%
  left_join(
    augment_metadata(phy) %>%
      select(sample_name, subject_id, site_group, swab_type2, Type),
    by = "sample_name"
  ) %>%
  mutate(filter = factor(filter, levels = filter_levels))

composition_stability_summary <- composition_stability %>%
  group_by(filter_id, filter) %>%
  summarise(
    n_samples = n(),
    mean_similarity = mean(composition_similarity),
    median_similarity = median(composition_similarity),
    minimum_similarity = min(composition_similarity),
    q25_similarity = quantile(composition_similarity, 0.25),
    .groups = "drop"
  )

write.csv(
  composition_stability,
  file.path(tab_dir, "sample_composition_stability.csv"),
  row.names = FALSE
)
write.csv(
  composition_stability_summary,
  file.path(tab_dir, "composition_stability_summary.csv"),
  row.names = FALSE
)

p_composition <- ggplot(
  composition_stability,
  aes(filter, composition_similarity * 100, fill = filter)
) +
  geom_boxplot(outlier.size = 0.6, linewidth = 0.35, show.legend = FALSE) +
  scale_y_continuous(limits = c(0, 101)) +
  labs(
    x = NULL,
    y = "Similarity to unfiltered composition (%)",
    title = "Sample-level community-composition stability"
  ) +
  theme_ms +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_plot(p_composition, "Figure5_Composition_stability", 9.2, 4.8)
composition_stability_summary
```

# ASV- and genus-level sharing across sampling systems

```{r shared-features}
presence_by_system <- function(ps) {
  otu <- otu_matrix(ps)
  md <- augment_metadata(ps)
  result <- sapply(paste0("Type ", 1:5), function(type_name) {
    ids <- md$sample_name[as.character(md$Type) == type_name]
    rowSums(otu[, ids, drop = FALSE] > 0) > 0
  })
  result <- as.data.frame(result, check.names = FALSE)
  rownames(result) <- rownames(otu)
  result
}

subset_site <- function(ps, site = NULL) {
  if (is.null(site) || length(site) == 0 || is.na(site)) return(ps)
  keep <- sample_names(ps)[as.character(sample_data(ps)$site_group) == site]
  prune_samples(keep, ps) %>% prune_taxa(taxa_sums(.) > 0, .)
}

summarise_feature_sharing <- function(ps, filter_id, rank_name, site = NULL) {
  ps_site <- subset_site(ps, site)
  if (rank_name == "Genus") {
    ps_site <- tax_glom(ps_site, taxrank = "Genus", NArm = FALSE) %>%
      prune_taxa(taxa_sums(.) > 0, .)
  }
  present <- presence_by_system(ps_site)
  n_systems <- rowSums(present)
  otu <- otu_matrix(ps_site)
  shared_ids <- rownames(present)[n_systems == 5]
  exclusive_ids <- rownames(present)[n_systems == 1]
  total_reads <- sum(otu)
  tibble(
    filter_id = filter_id,
    filter = filter_labels[[filter_id]],
    taxonomic_level = rank_name,
    subset = ifelse(is.null(site) || length(site) == 0 || is.na(site), "Total", site),
    retained_features = nrow(present),
    shared_by_all_five = length(shared_ids),
    shared_by_all_five_feature_pct = 100 * length(shared_ids) / nrow(present),
    shared_by_all_five_reads = sum(otu[shared_ids, , drop = FALSE]),
    shared_by_all_five_read_pct = 100 * sum(otu[shared_ids, , drop = FALSE]) / total_reads,
    system_exclusive_features = length(exclusive_ids),
    system_exclusive_reads = sum(otu[exclusive_ids, , drop = FALSE]),
    system_exclusive_read_pct = 100 * sum(otu[exclusive_ids, , drop = FALSE]) / total_reads
  )
}

shared_feature_summary <- bind_rows(lapply(filter_ids, function(id) {
  ps <- filtered_ps[[id]]
  bind_rows(lapply(c("ASV", "Genus"), function(rank_name) {
    bind_rows(lapply(c(NA_character_, "Antecubital", "Forehead"), function(site) {
      summarise_feature_sharing(ps, id, rank_name, site)
    }))
  }))
})) %>%
  mutate(
    filter = factor(filter, levels = filter_levels),
    taxonomic_level = factor(taxonomic_level, levels = c("ASV", "Genus")),
    subset = factor(subset, levels = c("Total", "Antecubital", "Forehead"))
  )

write.csv(
  shared_feature_summary %>%
    mutate(across(c(filter, taxonomic_level, subset), as.character)),
  file.path(tab_dir, "shared_feature_summary.csv"),
  row.names = FALSE
)

p_shared <- shared_feature_summary %>%
  ggplot(
    aes(filter, shared_by_all_five_read_pct, colour = taxonomic_level, group = taxonomic_level)
  ) +
  geom_line(linewidth = 0.65) +
  geom_point(size = 1.8) +
  facet_wrap(~subset, ncol = 1) +
  scale_colour_manual(values = metric_col[c("ASV", "Genus")]) +
  scale_y_continuous(limits = c(0, 101)) +
  labs(
    x = NULL,
    y = "Reads from features shared by all five systems (%)",
    colour = "Level",
    title = "Stability of universally shared bacterial features"
  ) +
  theme_ms +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_plot(p_shared, "Figure6_Shared_feature_sensitivity", 9.2, 8.3)
shared_feature_summary
```

# Sampling-system-exclusive features

```{r exclusive-features}
exclusive_feature_detail <- function(ps, filter_id, rank_name) {
  if (rank_name == "Genus") {
    ps <- tax_glom(ps, taxrank = "Genus", NArm = FALSE) %>%
      prune_taxa(taxa_sums(.) > 0, .)
  }
  present <- presence_by_system(ps)
  n_systems <- rowSums(present)
  exclusive_ids <- rownames(present)[n_systems == 1]
  if (length(exclusive_ids) == 0) return(tibble())
  otu <- otu_matrix(ps)
  md <- augment_metadata(ps)
  total_reads <- sum(otu)

  bind_rows(lapply(exclusive_ids, function(feature_id) {
    exclusive_type <- colnames(present)[which(as.logical(present[feature_id, , drop = TRUE]))]
    sample_ids <- md$sample_name[as.character(md$Type) == exclusive_type]
    subject_ids <- unique(md$subject_id[
      md$sample_name %in% sample_ids[otu[feature_id, sample_ids, drop = TRUE] > 0]
    ])
    tibble(
      filter_id = filter_id,
      filter = filter_labels[[filter_id]],
      taxonomic_level = rank_name,
      feature_id = feature_id,
      exclusive_type = exclusive_type,
      feature_reads = sum(otu[feature_id, ]),
      feature_read_pct = 100 * sum(otu[feature_id, ]) / total_reads,
      participant_prevalence = length(subject_ids)
    )
  }))
}

exclusive_detail <- bind_rows(lapply(filter_ids, function(id) {
  bind_rows(
    exclusive_feature_detail(filtered_ps[[id]], id, "ASV"),
    exclusive_feature_detail(filtered_ps[[id]], id, "Genus")
  )
}))

exclusive_summary <- exclusive_detail %>%
  group_by(filter_id, filter, taxonomic_level, exclusive_type) %>%
  summarise(
    exclusive_features = n(),
    exclusive_read_pct = sum(feature_read_pct),
    median_participant_prevalence = median(participant_prevalence),
    maximum_participant_prevalence = max(participant_prevalence),
    .groups = "drop"
  ) %>%
  mutate(
    filter = factor(filter, levels = filter_levels),
    exclusive_type = factor(exclusive_type, levels = paste0("Type ", 1:5))
  )

write.csv(exclusive_detail, file.path(tab_dir, "system_exclusive_feature_detail.csv"), row.names = FALSE)
write.csv(exclusive_summary, file.path(tab_dir, "system_exclusive_feature_summary.csv"), row.names = FALSE)

p_exclusive <- ggplot(
  exclusive_summary,
  aes(filter, exclusive_read_pct, fill = exclusive_type)
) +
  geom_col() +
  facet_wrap(~taxonomic_level, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = type_col, drop = FALSE) +
  labs(
    x = NULL,
    y = "Reads assigned to system-exclusive features (%)",
    fill = "Sampling system",
    title = "Sampling-system-exclusive bacterial features"
  ) +
  theme_ms +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_plot(p_exclusive, "Figure7_System_exclusive_features", 9.2, 7.2)
exclusive_summary
```

# Existing decontam threshold analysis

The threshold comparison below is imported from the completed prevalence-based
decontam sensitivity analysis. It is not recalculated here.

```{r decontam-summary}
decontam_path <- file.path(getwd(), "Tables", "Decontam_permanova_summary.txt")

if (file.exists(decontam_path)) {
  decontam_summary <- read.delim(
    decontam_path,
    row.names = 1,
    check.names = FALSE,
    stringsAsFactors = FALSE
  ) %>% as_tibble()

  write.csv(
    decontam_summary,
    file.path(tab_dir, "decontam_threshold_summary_existing_results.csv"),
    row.names = FALSE
  )

  p_decontam_reads <- decontam_summary %>%
    select(threshold, reads_retained_skin, reads_retained_ctrl) %>%
    pivot_longer(-threshold, names_to = "sample_class", values_to = "reads_retained_pct") %>%
    mutate(
      sample_class = recode(
        sample_class,
        reads_retained_skin = "True samples",
        reads_retained_ctrl = "Controls"
      )
    ) %>%
    ggplot(aes(threshold, reads_retained_pct, colour = sample_class)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.8) +
    scale_x_continuous(breaks = seq(0.1, 0.9, 0.1)) +
    labs(
      x = "decontam prevalence threshold",
      y = "Reads retained (%)",
      colour = "Sample class",
      title = "Read retention across decontam thresholds"
    ) +
    theme_ms

  p_decontam_asv <- ggplot(
    decontam_summary,
    aes(threshold, n_contam_ASV)
  ) +
    geom_line(colour = "#6a3d9a", linewidth = 0.7) +
    geom_point(colour = "#6a3d9a", size = 1.8) +
    scale_x_continuous(breaks = seq(0.1, 0.9, 0.1)) +
    labs(
      x = "decontam prevalence threshold",
      y = "ASVs classified as contaminants",
      title = "Contaminant calls across thresholds"
    ) +
    theme_ms

  p_decontam <- p_decontam_reads | p_decontam_asv
  save_plot(p_decontam, "Figure8_Decontam_threshold_sensitivity", 10, 4.4)
  decontam_summary
} else {
  warning("Existing decontam summary was not found: ", decontam_path)
}
```

# Integrated decision table

```{r decision-table}
system_permanova <- permanova_results %>%
  filter(variable_id == "swab_type2") %>%
  select(
    filter_id, distance_id,
    sampling_system_R2 = R2,
    sampling_system_p = p_value
  ) %>%
  pivot_wider(
    names_from = distance_id,
    values_from = c(sampling_system_R2, sampling_system_p),
    names_glue = "{.value}_{distance_id}"
  )

shared_total <- shared_feature_summary %>%
  filter(subset == "Total") %>%
  select(
    filter_id, taxonomic_level,
    shared_by_all_five, shared_by_all_five_read_pct,
    system_exclusive_features, system_exclusive_read_pct
  ) %>%
  pivot_wider(
    names_from = taxonomic_level,
    values_from = c(
      shared_by_all_five,
      shared_by_all_five_read_pct,
      system_exclusive_features,
      system_exclusive_read_pct
    ),
    names_glue = "{.value}_{taxonomic_level}"
  )

decision_summary <- retention_summary %>%
  mutate(filter = as.character(filter)) %>%
  left_join(
    composition_stability_summary %>%
      select(filter_id, median_composition_similarity = median_similarity),
    by = "filter_id"
  ) %>%
  left_join(system_permanova, by = "filter_id") %>%
  left_join(shared_total, by = "filter_id") %>%
  mutate(
    prevalence_filter_equivalent_to_ge5 = filter_id %in% c(
      "prevalence_ge_5", "prevalence_ge_10pct"
    ),
    interpretation = case_when(
      filter_id == "no_filter" ~ "Post-decontam reference condition",
      filter_id == "remove_count_le_1" ~ "Minimum rare-feature filter (singleton removal)",
      filter_id == "remove_count_le_2" ~ "Conservative count filter (counts <=2 removed)",
      filter_id == "remove_count_le_10" ~ "Stronger count filter; use as sensitivity analysis",
      filter_id == "prevalence_ge_2" ~ "Minimum prevalence filter",
      filter_id == "prevalence_ge_5" ~ "Prevalence in at least five of 50 true samples",
      filter_id == "prevalence_ge_10pct" ~ "Equivalent to prevalence >=5 for n=50",
      filter_id == "mean_abundance_ge_0.1pct" ~ "Dominant-community abundance filter",
      filter_id == "mean_abundance_ge_1pct" ~ "Aggressive dominant-community-only filter",
      TRUE ~ NA_character_
    )
  )

write.csv(
  decision_summary,
  file.path(tab_dir, "filtering_decision_summary.csv"),
  row.names = FALSE
)

writeLines(
  c(
    "Filtering sensitivity analysis outputs",
    paste0("Generated: ", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")),
    paste0("Starting object: Phyloseq/phy_F270R240_260812_v4.rds"),
    paste0("True samples: ", nsamples(phy)),
    paste0("Post-decontam ASVs before additional filtering: ", ntaxa(phy)),
    paste0("Post-decontam reads before additional filtering: ", sum(sample_sums(phy))),
    paste0("Fixed rarefaction depth: ", target_depth),
    "",
    "Important equivalence:",
    paste0(
      "With ", nsamples(phy), " true samples, prevalence >=10% requires detection in ",
      ceiling(0.10 * nsamples(phy)),
      " samples and is therefore identical to prevalence >=5 samples."
    ),
    "",
    "See filtering_decision_summary.csv for the integrated comparison."
  ),
  file.path(out_dir, "README_results.txt")
)

decision_summary
```

# Session information

```{r session-info}
writeLines(capture.output(sessionInfo()), file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
