---
title: "Integrated decontam, filtering, and sampling-system sensitivity analysis"
author: "DT Swab Study"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
project_library <- file.path(getwd(), "R_library")
if (dir.exists(project_library)) .libPaths(c(project_library, .libPaths()))
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Objective

The primary study question remains the comparison of five integrated swab-medium
sampling systems. This analysis tests whether that conclusion changes when
prevalence-based decontamination and rare-feature filtering decisions are varied.

Two prespecified sensitivity axes are evaluated:

1. No decontamination and decontam prevalence thresholds 0.1--0.9, with singleton
   removal applied to every condition.
2. Nine rare-feature filters at the selected decontam threshold 0.5.

The phylogenetic tree stored in the existing phyloseq objects is used without
modification. No rerooting, polytomy resolution, or branch-length transformation is
performed. All permutation tests use 9,999 permutations and random seed 42.

```{r packages}
suppressPackageStartupMessages({
  library(phyloseq)
  library(decontam)
  library(vegan)
  library(permute)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(patchwork)
  library(RColorBrewer)
})

set.seed(42)
permutations_n <- 9999L
target_depth <- 5876L

out_dir <- file.path(getwd(), "Integrated_sensitivity_output")
fig_dir <- file.path(out_dir, "Figures")
tab_dir <- file.path(out_dir, "Tables")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(tab_dir, recursive = TRUE, showWarnings = FALSE)

distance_methods <- c(
  weighted_unifrac = "wunifrac",
  unweighted_unifrac = "unifrac",
  bray_curtis = "bray"
)

thresholds <- seq(0.1, 0.9, by = 0.1)
threshold_labels <- c("None", sprintf("%0.1f", thresholds))

filter_ids <- c(
  "no_filter",
  "remove_count_le_1",
  "remove_count_le_2",
  "remove_count_le_10",
  "prevalence_ge_2",
  "prevalence_ge_5",
  "prevalence_ge_10pct",
  "mean_abundance_ge_0.1pct",
  "mean_abundance_ge_1pct"
)

filter_labels <- c(
  no_filter = "No additional filter",
  remove_count_le_1 = "Singleton removal",
  remove_count_le_2 = "Remove total count <=2",
  remove_count_le_10 = "Remove total count <=10",
  prevalence_ge_2 = "Prevalence >=2 samples",
  prevalence_ge_5 = "Prevalence >=5 samples",
  prevalence_ge_10pct = "Prevalence >=10% samples",
  `mean_abundance_ge_0.1pct` = "Mean abundance >=0.1%",
  `mean_abundance_ge_1pct` = "Mean abundance >=1%"
)

otu_matrix <- function(ps) {
  otu <- as(otu_table(ps), "matrix")
  if (!taxa_are_rows(ps)) otu <- t(otu)
  otu
}

prune_nonzero <- function(ps) {
  prune_taxa(taxa_sums(ps) > 0, ps)
}
```

# Reconstruct the original decontam conditions

```{r reconstruct-decontam}
phy_pre <- readRDS("Phyloseq/phy_F270R240_260812_v3.rds")
original_tree <- phy_tree(phy_pre)

# Restore control labels from the original metadata by exact sample ID. The five
# controls are recorded as negative-control in the metadata and were converted to
# control in the original preprocessing workflow before decontam was run.
metadata_source <- read.delim(
  "metadata/DT_metadata.tsv",
  check.names = FALSE,
  stringsAsFactors = FALSE
)
rownames(metadata_source) <- metadata_source$sample_id
stopifnot(all(sample_names(phy_pre) %in% rownames(metadata_source)))
metadata_source <- metadata_source[sample_names(phy_pre), , drop = FALSE]
metadata_source$control_status <- ifelse(
  metadata_source$control_status == "negative-control",
  "control",
  metadata_source$control_status
)
stopifnot(
  sum(metadata_source$control_status == "sample") == 50L,
  sum(metadata_source$control_status == "control") == 5L
)
sample_data(phy_pre) <- sample_data(metadata_source)
sample_data(phy_pre)$is.neg <- sample_data(phy_pre)$control_status == "control"

decontam_calls <- setNames(
  lapply(thresholds, function(threshold) {
    isContaminant(
      phy_pre,
      method = "prevalence",
      neg = "is.neg",
      threshold = threshold
    )
  }),
  sprintf("%0.1f", thresholds)
)

make_clean_all <- function(threshold_label) {
  if (threshold_label == "None") return(phy_pre)
  calls <- decontam_calls[[threshold_label]]
  prune_taxa(!calls$contaminant, phy_pre)
}

clean_all_objects <- setNames(
  lapply(threshold_labels, make_clean_all),
  threshold_labels
)

true_no_filter_objects <- lapply(clean_all_objects, function(ps) {
  ps <- subset_samples(ps, control_status == "sample")
  prune_nonzero(ps)
})

control_objects <- lapply(clean_all_objects, function(ps) {
  ps <- subset_samples(ps, control_status == "control")
  prune_nonzero(ps)
})

# The threshold axis uses the same singleton definition in every condition.
threshold_objects <- lapply(true_no_filter_objects, function(ps) {
  prune_taxa(taxa_sums(ps) > 1, ps) %>% prune_nonzero()
})

apply_feature_filter <- function(ps, filter_id) {
  otu <- otu_matrix(ps)
  total_count <- rowSums(otu)
  prevalence <- rowSums(otu > 0)
  rel <- sweep(otu, 2, colSums(otu), "/")
  rel[!is.finite(rel)] <- 0
  mean_rel_pct <- rowMeans(rel) * 100

  keep <- switch(
    filter_id,
    no_filter = total_count > 0,
    remove_count_le_1 = total_count > 1,
    remove_count_le_2 = total_count > 2,
    remove_count_le_10 = total_count > 10,
    prevalence_ge_2 = prevalence >= 2,
    prevalence_ge_5 = prevalence >= 5,
    prevalence_ge_10pct = prevalence >= ceiling(0.10 * nsamples(ps)),
    `mean_abundance_ge_0.1pct` = mean_rel_pct >= 0.1,
    `mean_abundance_ge_1pct` = mean_rel_pct >= 1,
    stop("Unknown filter: ", filter_id)
  )
  prune_taxa(keep, ps) %>% prune_nonzero()
}

filter_base <- true_no_filter_objects[["0.5"]]
filter_objects <- setNames(
  lapply(filter_ids, function(filter_id) {
    apply_feature_filter(filter_base, filter_id)
  }),
  filter_ids
)
```

# Validation of reconstruction and tree preservation

```{r reconstruction-validation}
phy_v4 <- readRDS("Phyloseq/phy_F270R240_260812_v4.rds")
phy_v4_true <- subset_samples(phy_v4, control_status == "sample") %>% prune_nonzero()
reconstructed_v05 <- true_no_filter_objects[["0.5"]]

validation <- tibble(
  check = c(
    "50 true and five control samples restored",
    "threshold 0.5 ASV count matches saved v4",
    "threshold 0.5 true-sample reads match saved v4",
    "threshold 0.5 ASV identities match saved v4",
    "preprocessing tree object was not modified",
    "prevalence >=10% equals prevalence >=5 at n=50"
  ),
  passed = c(
    sum(sample_data(phy_pre)$control_status == "sample") == 50L &&
      sum(sample_data(phy_pre)$control_status == "control") == 5L,
    ntaxa(reconstructed_v05) == ntaxa(phy_v4_true),
    sum(sample_sums(reconstructed_v05)) == sum(sample_sums(phy_v4_true)),
    setequal(taxa_names(reconstructed_v05), taxa_names(phy_v4_true)),
    identical(phy_tree(phy_pre), original_tree),
    setequal(
      taxa_names(filter_objects[["prevalence_ge_10pct"]]),
      taxa_names(filter_objects[["prevalence_ge_5"]])
    )
  ),
  observed = c(
    paste(sum(sample_data(phy_pre)$control_status == "sample"),
          sum(sample_data(phy_pre)$control_status == "control"), sep = " / "),
    paste(ntaxa(reconstructed_v05), ntaxa(phy_v4_true), sep = " vs "),
    paste(sum(sample_sums(reconstructed_v05)), sum(sample_sums(phy_v4_true)), sep = " vs "),
    as.character(setequal(taxa_names(reconstructed_v05), taxa_names(phy_v4_true))),
    as.character(identical(phy_tree(phy_pre), original_tree)),
    as.character(setequal(
      taxa_names(filter_objects[["prevalence_ge_10pct"]]),
      taxa_names(filter_objects[["prevalence_ge_5"]])
    ))
  )
)
write.csv(validation, file.path(tab_dir, "Table00_validation_sanity_checks.csv"), row.names = FALSE)
if (!all(validation$passed)) stop("Integrated analysis reconstruction failed")
validation
```

# Data retention and sample-level read removal

```{r retention-removal}
pre_true <- true_no_filter_objects[["None"]]
pre_control <- control_objects[["None"]]
pre_true_reads <- sum(sample_sums(pre_true))
pre_control_reads <- sum(sample_sums(pre_control))

threshold_retention <- bind_rows(lapply(threshold_labels, function(label) {
  true_ps <- true_no_filter_objects[[label]]
  threshold_ps <- threshold_objects[[label]]
  ctrl_ps <- control_objects[[label]]
  tibble(
    threshold = label,
    threshold_numeric = ifelse(label == "None", 0, as.numeric(label)),
    contaminant_asvs = ifelse(
      label == "None", 0L, sum(decontam_calls[[label]]$contaminant)
    ),
    true_asvs_before_singleton = ntaxa(true_ps),
    true_asvs_after_singleton = ntaxa(threshold_ps),
    true_reads = sum(sample_sums(true_ps)),
    true_reads_retained_pct = 100 * sum(sample_sums(true_ps)) / pre_true_reads,
    control_reads = sum(sample_sums(ctrl_ps)),
    control_reads_retained_pct = 100 * sum(sample_sums(ctrl_ps)) / pre_control_reads,
    samples_below_5876 = sum(sample_sums(threshold_ps) < target_depth),
    samples_available_at_5876 = sum(sample_sums(threshold_ps) >= target_depth),
    minimum_true_sample_depth = min(sample_sums(threshold_ps))
  )
}))

filter_retention <- bind_rows(lapply(filter_ids, function(filter_id) {
  ps <- filter_objects[[filter_id]]
  tibble(
    filter_id = filter_id,
    filter = unname(filter_labels[filter_id]),
    retained_asvs = ntaxa(ps),
    retained_reads = sum(sample_sums(ps)),
    reads_retained_vs_v05_pct = 100 * sum(sample_sums(ps)) / sum(sample_sums(filter_base)),
    samples_below_5876 = sum(sample_sums(ps) < target_depth),
    samples_available_at_5876 = sum(sample_sums(ps) >= target_depth),
    minimum_depth = min(sample_sums(ps))
  )
}))

pre_sample_reads <- sample_sums(phy_pre)
sample_removal <- bind_rows(lapply(threshold_labels, function(label) {
  post_reads <- sample_sums(clean_all_objects[[label]])
  md <- data.frame(sample_data(phy_pre), check.names = FALSE)
  tibble(
    sample_name = names(pre_sample_reads),
    threshold = label,
    threshold_numeric = ifelse(label == "None", 0, as.numeric(label)),
    reads_before = as.numeric(pre_sample_reads),
    reads_after = as.numeric(post_reads[names(pre_sample_reads)]),
    reads_removed = reads_before - reads_after,
    removed_pct = 100 * reads_removed / reads_before,
    subject_id = as.character(md[names(pre_sample_reads), "subject_id"]),
    site_group = as.character(md[names(pre_sample_reads), "site_group"]),
    swab_type2 = as.character(md[names(pre_sample_reads), "swab_type2"]),
    control_status = as.character(md[names(pre_sample_reads), "control_status"])
  )
}))

write.csv(threshold_retention, file.path(tab_dir, "Table01_decontam_threshold_retention.csv"), row.names = FALSE)
write.csv(filter_retention, file.path(tab_dir, "Table02_filter_retention_at_threshold_0.5.csv"), row.names = FALSE)
write.csv(sample_removal, file.path(tab_dir, "Table03_sample_level_read_removal.csv"), row.names = FALSE)
```

# Repeated-measures comparison of removal by sampling system

```{r removal-repeated-measures}
run_removal_friedman <- function(threshold_label, site) {
  dat <- sample_removal %>%
    filter(
      threshold == threshold_label,
      control_status == "sample",
      site_group == site
    )
  fit <- tryCatch(
    friedman.test(
      y = dat$removed_pct,
      groups = dat$swab_type2,
      blocks = dat$subject_id
    ),
    error = function(e) NULL
  )
  tibble(
    threshold = threshold_label,
    site_group = site,
    n_participants = n_distinct(dat$subject_id),
    n_systems = n_distinct(dat$swab_type2),
    statistic = if (is.null(fit)) NA_real_ else unname(fit$statistic),
    df = if (is.null(fit)) NA_real_ else unname(fit$parameter),
    p_value = if (is.null(fit)) NA_real_ else fit$p.value,
    test = "Friedman repeated-measures omnibus test"
  )
}

removal_friedman <- bind_rows(lapply(sprintf("%0.1f", thresholds), function(label) {
  bind_rows(lapply(c("Antecubital", "Forehead"), function(site) {
    run_removal_friedman(label, site)
  }))
})) %>%
  group_by(site_group) %>%
  mutate(
    p_adjusted_BH = p.adjust(p_value, method = "BH")
  ) %>%
  ungroup()

write.csv(removal_friedman, file.path(tab_dir, "Table04_sampling_system_read_removal_Friedman.csv"), row.names = FALSE)
```

# Rarefaction and community statistics

```{r community-functions}
rarefy_at_target <- function(ps) {
  keep <- sample_names(ps)[sample_sums(ps) >= target_depth]
  if (length(keep) < 2L) return(NULL)
  ps <- prune_samples(keep, ps) %>% prune_nonzero()
  set.seed(42)
  rarefy_even_depth(
    ps,
    sample.size = target_depth,
    rngseed = 42,
    replace = FALSE,
    trimOTUs = TRUE,
    verbose = FALSE
  )
}

threshold_rarefied <- lapply(threshold_objects, rarefy_at_target)
filter_rarefied <- lapply(filter_objects, rarefy_at_target)

run_permanova <- function(ps, condition_axis, condition_id, distance_name,
                          distance_method, variable) {
  if (is.null(ps)) return(tibble())
  dist_obj <- phyloseq::distance(ps, method = distance_method)
  md <- data.frame(sample_data(ps), check.names = FALSE)
  model_formula <- as.formula(paste("dist_obj ~", variable))
  blocked <- variable != "subject_id"
  set.seed(42)
  fit <- if (blocked) {
    adonis2(model_formula, data = md, permutations = permutations_n,
            strata = md$subject_id)
  } else {
    adonis2(model_formula, data = md, permutations = permutations_n)
  }
  tibble(
    condition_axis = condition_axis,
    condition_id = condition_id,
    distance = distance_name,
    variable = variable,
    n_samples = nrow(md),
    n_participants = n_distinct(md$subject_id),
    df = fit$Df[1],
    pseudo_F = fit$F[1],
    R2 = fit$R2[1],
    p_value = fit$`Pr(>F)`[1],
    permutations = permutations_n,
    permutation_scheme = ifelse(blocked, "restricted within participant", "unrestricted")
  )
}

run_permdisp <- function(ps, condition_axis, condition_id, distance_name,
                         distance_method) {
  if (is.null(ps)) return(tibble())
  dist_obj <- phyloseq::distance(ps, method = distance_method)
  md <- data.frame(sample_data(ps), check.names = FALSE)
  dispersion <- betadisper(dist_obj, md$swab_type2, type = "median", bias.adjust = TRUE)
  design <- permute::how(nperm = permutations_n, blocks = md$subject_id)
  set.seed(42)
  fit <- permutest(dispersion, permutations = design)
  tibble(
    condition_axis = condition_axis,
    condition_id = condition_id,
    distance = distance_name,
    n_samples = nrow(md),
    F = fit$tab$F[1],
    p_value = fit$tab$`Pr(>F)`[1],
    permutations = permutations_n,
    permutation_scheme = "restricted within participant"
  )
}

run_condition_statistics <- function(object_list, condition_axis) {
  permanova <- bind_rows(lapply(names(object_list), function(condition_id) {
    ps <- object_list[[condition_id]]
    bind_rows(lapply(names(distance_methods), function(distance_name) {
      bind_rows(lapply(c("subject_id", "site_group", "swab_type2"), function(variable) {
        run_permanova(
          ps, condition_axis, condition_id,
          distance_name, distance_methods[[distance_name]], variable
        )
      }))
    }))
  }))

  permdisp <- bind_rows(lapply(names(object_list), function(condition_id) {
    ps <- object_list[[condition_id]]
    bind_rows(lapply(names(distance_methods), function(distance_name) {
      run_permdisp(
        ps, condition_axis, condition_id,
        distance_name, distance_methods[[distance_name]]
      )
    }))
  }))
  list(permanova = permanova, permdisp = permdisp)
}

threshold_statistics <- run_condition_statistics(threshold_rarefied, "decontam_threshold")
filter_statistics <- run_condition_statistics(filter_rarefied, "filter_at_threshold_0.5")

permanova_all <- bind_rows(
  threshold_statistics$permanova,
  filter_statistics$permanova
)
permdisp_all <- bind_rows(
  threshold_statistics$permdisp,
  filter_statistics$permdisp
)

write.csv(permanova_all, file.path(tab_dir, "Table05_PERMANOVA_all_conditions.csv"), row.names = FALSE)
write.csv(permdisp_all, file.path(tab_dir, "Table06_PERMDISP_all_conditions.csv"), row.names = FALSE)
```

# Alpha diversity with repeated-measures omnibus tests

```{r alpha-diversity}
run_alpha_friedman <- function(ps, condition_axis, condition_id) {
  if (is.null(ps)) return(tibble())
  md <- data.frame(sample_data(ps), check.names = FALSE) %>%
    rownames_to_column("sample_name") %>%
    mutate(sample_name = sub("^X", "", sample_name))
  alpha <- estimate_richness(ps, measures = c("Shannon", "Observed")) %>%
    rownames_to_column("sample_name") %>%
    mutate(sample_name = sub("^X", "", sample_name)) %>%
    left_join(md, by = "sample_name")

  bind_rows(lapply(unique(as.character(alpha$site_group)), function(site) {
    bind_rows(lapply(c("Shannon", "Observed"), function(response) {
      dat <- alpha %>% filter(site_group == site)
      complete_design <- nrow(dat) == 25L &&
        n_distinct(dat$subject_id) == 5L &&
        n_distinct(dat$swab_type2) == 5L
      fit <- if (complete_design) {
        friedman.test(
          y = dat[[response]],
          groups = dat$swab_type2,
          blocks = dat$subject_id
        )
      } else NULL
      tibble(
        condition_axis = condition_axis,
        condition_id = condition_id,
        site_group = site,
        response = response,
        n_samples = nrow(dat),
        n_participants = n_distinct(dat$subject_id),
        statistic = if (is.null(fit)) NA_real_ else unname(fit$statistic),
        df = if (is.null(fit)) NA_real_ else unname(fit$parameter),
        p_value = if (is.null(fit)) NA_real_ else fit$p.value,
        complete_5_by_5_design = complete_design
      )
    }))
  }))
}

alpha_results <- bind_rows(
  bind_rows(lapply(names(threshold_rarefied), function(id) {
    run_alpha_friedman(threshold_rarefied[[id]], "decontam_threshold", id)
  })),
  bind_rows(lapply(names(filter_rarefied), function(id) {
    run_alpha_friedman(filter_rarefied[[id]], "filter_at_threshold_0.5", id)
  }))
)

write.csv(alpha_results, file.path(tab_dir, "Table07_alpha_diversity_Friedman.csv"), row.names = FALSE)
```

# Composition similarity, shared features, and matched-block agreement

```{r composition-sharing-functions}
composition_similarity <- function(reference, candidate) {
  shared_samples <- intersect(sample_names(reference), sample_names(candidate))
  ref <- prune_samples(shared_samples, reference)
  cand <- prune_samples(shared_samples, candidate)
  all_taxa <- union(taxa_names(ref), taxa_names(cand))

  aligned_relative <- function(ps) {
    otu <- otu_matrix(ps)
    aligned <- matrix(
      0,
      nrow = length(all_taxa),
      ncol = length(shared_samples),
      dimnames = list(all_taxa, shared_samples)
    )
    aligned[rownames(otu), colnames(otu)] <- otu
    rel <- sweep(aligned, 2, colSums(aligned), "/")
    rel[!is.finite(rel)] <- 0
    rel
  }

  ref_rel <- aligned_relative(ref)
  cand_rel <- aligned_relative(cand)
  tibble(
    sample_name = shared_samples,
    similarity = vapply(shared_samples, function(sample_name) {
      1 - sum(abs(ref_rel[, sample_name] - cand_rel[, sample_name])) / 2
    }, numeric(1))
  )
}

prepare_rank_object <- function(ps, rank_name) {
  if (rank_name == "ASV") return(ps)
  tax <- as(tax_table(ps), "matrix")
  genus <- as.character(tax[, "Genus"])
  genus[is.na(genus) | genus == ""] <- "Unclassified_genus"
  tax[, "Genus"] <- genus
  tax_table(ps) <- tax_table(tax)
  tax_glom(ps, taxrank = "Genus", NArm = FALSE)
}

aggregate_shared_summary <- function(ps, rank_name) {
  work <- prepare_rank_object(ps, rank_name)
  otu <- otu_matrix(work)
  md <- data.frame(sample_data(work), check.names = FALSE)
  systems <- sort(unique(as.character(md$swab_type2)))
  present <- sapply(systems, function(system) {
    samples <- rownames(md)[md$swab_type2 == system]
    rowSums(otu[, samples, drop = FALSE]) > 0
  })
  shared <- rownames(otu)[rowSums(present) == length(systems)]
  exclusive <- rownames(otu)[rowSums(present) == 1L]
  tibble(
    taxonomic_level = rank_name,
    total_features = nrow(otu),
    shared_all_systems_features = length(shared),
    shared_all_systems_read_pct = 100 * sum(otu[shared, , drop = FALSE]) / sum(otu),
    system_exclusive_features = length(exclusive),
    system_exclusive_read_pct = 100 * sum(otu[exclusive, , drop = FALSE]) / sum(otu)
  )
}

matched_block_summary <- function(ps, rank_name) {
  work <- prepare_rank_object(ps, rank_name)
  otu <- otu_matrix(work)
  md <- data.frame(sample_data(work), check.names = FALSE)
  blocks <- md %>% rownames_to_column("sample_name") %>%
    distinct(subject_id, site_group)

  bind_rows(lapply(seq_len(nrow(blocks)), function(i) {
    block_md <- md[
      md$subject_id == blocks$subject_id[i] &
        md$site_group == blocks$site_group[i],
      , drop = FALSE
    ]
    samples <- rownames(block_md)
    block_otu <- otu[, samples, drop = FALSE]
    present <- block_otu > 0
    shared <- rownames(block_otu)[rowSums(present) == ncol(block_otu)]
    union_ids <- rownames(block_otu)[rowSums(present) > 0]
    tibble(
      taxonomic_level = rank_name,
      subject_id = blocks$subject_id[i],
      site_group = blocks$site_group[i],
      systems_present = n_distinct(block_md$swab_type2),
      union_features = length(union_ids),
      shared_all_five_features = length(shared),
      intersection_over_union = length(shared) / length(union_ids),
      shared_all_five_read_pct = 100 * sum(block_otu[shared, , drop = FALSE]) / sum(block_otu)
    )
  }))
}

summarise_condition_composition <- function(object_list, condition_axis, reference) {
  similarity <- bind_rows(lapply(names(object_list), function(id) {
    composition_similarity(reference, object_list[[id]]) %>%
      summarise(
        mean_similarity = mean(similarity),
        median_similarity = median(similarity),
        minimum_similarity = min(similarity)
      ) %>%
      mutate(condition_axis = condition_axis, condition_id = id, .before = 1)
  }))

  aggregate <- bind_rows(lapply(names(object_list), function(id) {
    bind_rows(
      aggregate_shared_summary(object_list[[id]], "ASV"),
      aggregate_shared_summary(object_list[[id]], "Genus")
    ) %>% mutate(condition_axis = condition_axis, condition_id = id, .before = 1)
  }))

  matched <- bind_rows(lapply(names(object_list), function(id) {
    bind_rows(
      matched_block_summary(object_list[[id]], "ASV"),
      matched_block_summary(object_list[[id]], "Genus")
    ) %>% mutate(condition_axis = condition_axis, condition_id = id, .before = 1)
  }))

  list(similarity = similarity, aggregate = aggregate, matched = matched)
}

reference_object <- threshold_objects[["None"]]
threshold_composition <- summarise_condition_composition(
  threshold_objects, "decontam_threshold", reference_object
)
filter_composition <- summarise_condition_composition(
  filter_objects, "filter_at_threshold_0.5", reference_object
)

composition_similarity_all <- bind_rows(
  threshold_composition$similarity,
  filter_composition$similarity
)
aggregate_sharing_all <- bind_rows(
  threshold_composition$aggregate,
  filter_composition$aggregate
)
matched_sharing_all <- bind_rows(
  threshold_composition$matched,
  filter_composition$matched
)

write.csv(composition_similarity_all, file.path(tab_dir, "Table08_composition_similarity.csv"), row.names = FALSE)
write.csv(aggregate_sharing_all, file.path(tab_dir, "Table09_aggregate_shared_features.csv"), row.names = FALSE)
write.csv(matched_sharing_all, file.path(tab_dir, "Table10_matched_block_shared_features.csv"), row.names = FALSE)
```

# Ordination concordance before and after decontam

```{r procrustes}
run_procrustes <- function(reference_ps, candidate_ps, condition_axis,
                           condition_id, reference_id,
                           distance_name, distance_method) {
  shared_samples <- intersect(sample_names(reference_ps), sample_names(candidate_ps))
  ref <- prune_samples(shared_samples, reference_ps) %>% prune_nonzero()
  cand <- prune_samples(shared_samples, candidate_ps) %>% prune_nonzero()

  ref_dist <- phyloseq::distance(ref, method = distance_method)
  cand_dist <- phyloseq::distance(cand, method = distance_method)
  ref_coord <- cmdscale(ref_dist, k = 2, eig = FALSE)
  cand_coord <- cmdscale(cand_dist, k = 2, eig = FALSE)
  cand_coord <- cand_coord[rownames(ref_coord), , drop = FALSE]
  md <- data.frame(sample_data(ref), check.names = FALSE)
  md <- md[rownames(ref_coord), , drop = FALSE]

  if (condition_id == reference_id) {
    return(tibble(
      condition_axis = condition_axis,
      condition_id = condition_id,
      distance = distance_name,
      n_samples = nrow(ref_coord),
      procrustes_correlation = 1,
      protest_p_value = NA_real_,
      permutations = permutations_n,
      permutation_scheme = "not applicable: self-comparison"
    ))
  }

  design <- permute::how(nperm = permutations_n, blocks = md$subject_id)
  set.seed(42)
  fit <- protest(ref_coord, cand_coord, permutations = design)
  tibble(
    condition_axis = condition_axis,
    condition_id = condition_id,
    distance = distance_name,
    n_samples = nrow(ref_coord),
    procrustes_correlation = unname(fit$t0),
    protest_p_value = fit$signif,
    permutations = permutations_n,
    permutation_scheme = "restricted within participant"
  )
}

threshold_procrustes <- bind_rows(lapply(names(threshold_rarefied), function(label) {
  candidate <- threshold_rarefied[[label]]
  bind_rows(lapply(names(distance_methods), function(distance_name) {
    run_procrustes(
      threshold_rarefied[["None"]],
      candidate,
      "decontam_threshold",
      label,
      "None",
      distance_name,
      distance_methods[[distance_name]]
    )
  }))
}))

filter_procrustes <- bind_rows(lapply(names(filter_rarefied), function(label) {
  candidate <- filter_rarefied[[label]]
  bind_rows(lapply(names(distance_methods), function(distance_name) {
    run_procrustes(
      filter_rarefied[["no_filter"]],
      candidate,
      "filter_at_threshold_0.5",
      label,
      "no_filter",
      distance_name,
      distance_methods[[distance_name]]
    )
  }))
}))

procrustes_results <- bind_rows(threshold_procrustes, filter_procrustes)

write.csv(procrustes_results, file.path(tab_dir, "Table11_ordination_Procrustes.csv"), row.names = FALSE)
```

# Fixed-sample matched-distance validation of filtering effects

This analysis directly tests whether filtering reduces dissimilarity among the
five sampling systems within the same participant and anatomical site. All 50
biological samples are retained for every filtering condition, so changes in
distance cannot be attributed to sample exclusion at the rarefaction step.
Distances between participants and between anatomical sites are calculated in
parallel to determine whether filtering simply compresses all biological
variation.

```{r fixed-sample-matched-distances}
filter_distance_objects <- lapply(filter_objects, function(ps) {
  stopifnot(setequal(sample_names(ps), sample_names(filter_base)))
  stopifnot(all(sample_sums(ps) > 0))
  transform_sample_counts(ps, function(x) x / sum(x)) %>% prune_nonzero()
})

extract_matched_distance_units <- function(ps, condition_id, distance_name,
                                           distance_method) {
  dist_obj <- phyloseq::distance(ps, method = distance_method)
  dist_mat <- as.matrix(dist_obj)
  md <- data.frame(sample_data(ps), check.names = FALSE) %>%
    rownames_to_column("sample_name")
  pairs <- which(upper.tri(dist_mat), arr.ind = TRUE)

  pair_df <- tibble(
    sample_1 = rownames(dist_mat)[pairs[, 1]],
    sample_2 = colnames(dist_mat)[pairs[, 2]],
    distance_value = dist_mat[pairs]
  ) %>%
    left_join(
      md %>% select(sample_1 = sample_name,
                    subject_1 = subject_id, site_1 = site_group,
                    system_1 = swab_type2),
      by = "sample_1"
    ) %>%
    left_join(
      md %>% select(sample_2 = sample_name,
                    subject_2 = subject_id, site_2 = site_group,
                    system_2 = swab_type2),
      by = "sample_2"
    ) %>%
    mutate(
      pair_class = case_when(
        subject_1 == subject_2 & site_1 == site_2 & system_1 != system_2 ~
          "within_participant_site_between_systems",
        subject_1 != subject_2 & site_1 == site_2 & system_1 == system_2 ~
          "between_participants_matched_site_system",
        subject_1 == subject_2 & site_1 != site_2 & system_1 == system_2 ~
          "between_sites_matched_participant_system",
        TRUE ~ NA_character_
      ),
      unit_id = case_when(
        pair_class == "within_participant_site_between_systems" ~
          paste(subject_1, site_1, sep = "__"),
        pair_class == "between_participants_matched_site_system" ~
          paste(site_1, system_1, sep = "__"),
        pair_class == "between_sites_matched_participant_system" ~
          paste(subject_1, system_1, sep = "__"),
        TRUE ~ NA_character_
      )
    ) %>%
    filter(!is.na(pair_class)) %>%
    group_by(pair_class, unit_id) %>%
    summarise(
      unit_median_distance = median(distance_value),
      pair_count = n(),
      .groups = "drop"
    ) %>%
    mutate(
      condition_axis = "filter_at_threshold_0.5_fixed_50_samples",
      condition_id = condition_id,
      distance = distance_name,
      .before = 1
    )
  pair_df
}

fixed_sample_distance_units <- bind_rows(lapply(names(filter_distance_objects), function(id) {
  bind_rows(lapply(names(distance_methods), function(distance_name) {
    extract_matched_distance_units(
      filter_distance_objects[[id]], id, distance_name,
      distance_methods[[distance_name]]
    )
  }))
}))

baseline_units <- fixed_sample_distance_units %>%
  filter(condition_id == "no_filter") %>%
  select(distance, pair_class, unit_id,
         baseline_distance = unit_median_distance)

fixed_sample_distance_units <- fixed_sample_distance_units %>%
  left_join(baseline_units, by = c("distance", "pair_class", "unit_id")) %>%
  mutate(
    absolute_change_from_no_filter = unit_median_distance - baseline_distance,
    percent_change_from_no_filter = ifelse(
      baseline_distance > 0,
      100 * absolute_change_from_no_filter / baseline_distance,
      NA_real_
    )
  )

fixed_sample_distance_summary <- fixed_sample_distance_units %>%
  group_by(condition_id, distance, pair_class) %>%
  summarise(
    n_units = n(),
    median_distance = median(unit_median_distance),
    IQR_distance = IQR(unit_median_distance),
    median_absolute_change = median(absolute_change_from_no_filter),
    median_percent_change = median(percent_change_from_no_filter, na.rm = TRUE),
    .groups = "drop"
  )

paired_filter_tests <- fixed_sample_distance_units %>%
  filter(condition_id != "no_filter") %>%
  group_by(condition_id, distance, pair_class) %>%
  group_modify(~{
    fit <- tryCatch(
      wilcox.test(
        .x$unit_median_distance,
        .x$baseline_distance,
        paired = TRUE,
        alternative = "two.sided",
        exact = FALSE
      ),
      error = function(e) NULL
    )
    tibble(
      n_units = nrow(.x),
      median_change = median(.x$absolute_change_from_no_filter),
      median_percent_change = median(.x$percent_change_from_no_filter, na.rm = TRUE),
      statistic = if (is.null(fit)) NA_real_ else unname(fit$statistic),
      p_value = if (is.null(fit)) NA_real_ else fit$p.value
    )
  }) %>%
  ungroup() %>%
  group_by(distance, pair_class) %>%
  mutate(
    p_adjusted_BH = p.adjust(p_value, method = "BH")
  ) %>%
  ungroup()

distance_signal_ratios <- fixed_sample_distance_summary %>%
  select(condition_id, distance, pair_class, median_distance) %>%
  pivot_wider(names_from = pair_class, values_from = median_distance) %>%
  mutate(
    participant_to_system_ratio =
      between_participants_matched_site_system /
      within_participant_site_between_systems,
    site_to_system_ratio =
      between_sites_matched_participant_system /
      within_participant_site_between_systems
  ) %>%
  group_by(distance) %>%
  mutate(
    participant_ratio_retained_pct = 100 * participant_to_system_ratio /
      participant_to_system_ratio[condition_id == "no_filter"],
    site_ratio_retained_pct = 100 * site_to_system_ratio /
      site_to_system_ratio[condition_id == "no_filter"]
  ) %>%
  ungroup()

write.csv(
  fixed_sample_distance_units,
  file.path(tab_dir, "Table17_fixed50_matched_distance_units.csv"),
  row.names = FALSE
)
write.csv(
  fixed_sample_distance_summary,
  file.path(tab_dir, "Table18_fixed50_matched_distance_summary.csv"),
  row.names = FALSE
)
write.csv(
  paired_filter_tests,
  file.path(tab_dir, "Table19_fixed50_paired_filter_tests.csv"),
  row.names = FALSE
)
write.csv(
  distance_signal_ratios,
  file.path(tab_dir, "Table20_fixed50_signal_ratios.csv"),
  row.names = FALSE
)
```

# Taxa removed at threshold 0.5

```{r removed-taxa}
calls_05 <- decontam_calls[["0.5"]]
removed_ids <- rownames(calls_05)[calls_05$contaminant]
otu_pre <- otu_matrix(phy_pre)
md_pre <- data.frame(sample_data(phy_pre), check.names = FALSE)
true_samples <- rownames(md_pre)[md_pre$control_status == "sample"]
control_samples <- rownames(md_pre)[md_pre$control_status == "control"]
tax <- as(tax_table(phy_pre), "matrix")

removed_taxa <- tibble(
  ASV = removed_ids,
  Phylum = as.character(tax[removed_ids, "Phylum"]),
  Genus = as.character(tax[removed_ids, "Genus"]),
  Species = as.character(tax[removed_ids, "Species"]),
  true_sample_reads = rowSums(otu_pre[removed_ids, true_samples, drop = FALSE]),
  control_reads = rowSums(otu_pre[removed_ids, control_samples, drop = FALSE]),
  total_reads = true_sample_reads + control_reads,
  true_sample_prevalence = rowSums(otu_pre[removed_ids, true_samples, drop = FALSE] > 0),
  control_prevalence = rowSums(otu_pre[removed_ids, control_samples, drop = FALSE] > 0)
) %>%
  arrange(desc(total_reads))

removed_genus <- removed_taxa %>%
  mutate(Genus = ifelse(is.na(Genus) | Genus == "", "Unclassified genus", Genus)) %>%
  group_by(Genus) %>%
  summarise(
    removed_asvs = n(),
    true_sample_reads = sum(true_sample_reads),
    control_reads = sum(control_reads),
    total_reads = sum(total_reads),
    .groups = "drop"
  ) %>%
  arrange(desc(total_reads))

write.csv(removed_taxa, file.path(tab_dir, "Table12_ASVs_removed_at_threshold_0.5.csv"), row.names = FALSE)
write.csv(removed_genus, file.path(tab_dir, "Table13_Genera_removed_at_threshold_0.5.csv"), row.names = FALSE)
```

# Integrated summaries

```{r integrated-summary}
threshold_effect_wide <- permanova_all %>%
  filter(condition_axis == "decontam_threshold") %>%
  select(condition_id, distance, variable, R2, p_value, n_samples) %>%
  pivot_wider(names_from = variable, values_from = c(R2, p_value))

filter_effect_wide <- permanova_all %>%
  filter(condition_axis == "filter_at_threshold_0.5") %>%
  select(condition_id, distance, variable, R2, p_value, n_samples) %>%
  pivot_wider(names_from = variable, values_from = c(R2, p_value))

threshold_integrated <- threshold_retention %>%
  rename(condition_id = threshold) %>%
  mutate(condition_axis = "decontam_threshold") %>%
  left_join(threshold_effect_wide, by = "condition_id") %>%
  left_join(
    permdisp_all %>%
      filter(condition_axis == "decontam_threshold") %>%
      select(condition_id, distance, system_permdisp_p = p_value),
    by = c("condition_id", "distance")
  ) %>%
  left_join(
    composition_similarity_all %>%
      filter(condition_axis == "decontam_threshold"),
    by = c("condition_id", "condition_axis")
  ) %>%
  left_join(
    aggregate_sharing_all %>%
      filter(condition_axis == "decontam_threshold", taxonomic_level == "Genus") %>%
      select(condition_id, distance_dummy = taxonomic_level,
             shared_genus_read_pct = shared_all_systems_read_pct,
             exclusive_genus_read_pct = system_exclusive_read_pct),
    by = "condition_id"
  ) %>%
  select(-distance_dummy)

filter_integrated <- filter_retention %>%
  rename(condition_id = filter_id) %>%
  left_join(filter_effect_wide, by = "condition_id") %>%
  left_join(
    permdisp_all %>%
      filter(condition_axis == "filter_at_threshold_0.5") %>%
      select(condition_id, distance, system_permdisp_p = p_value),
    by = c("condition_id", "distance")
  ) %>%
  left_join(
    composition_similarity_all %>%
      filter(condition_axis == "filter_at_threshold_0.5"),
    by = "condition_id"
  ) %>%
  left_join(
    aggregate_sharing_all %>%
      filter(condition_axis == "filter_at_threshold_0.5", taxonomic_level == "Genus") %>%
      select(condition_id,
             shared_genus_read_pct = shared_all_systems_read_pct,
             exclusive_genus_read_pct = system_exclusive_read_pct),
    by = "condition_id"
  )

write.csv(threshold_integrated, file.path(tab_dir, "Table14_integrated_decontam_threshold_results.csv"), row.names = FALSE)
write.csv(filter_integrated, file.path(tab_dir, "Table15_integrated_filter_results.csv"), row.names = FALSE)
```

# Figures

```{r figures, fig.width=12, fig.height=8}
threshold_plot_data <- threshold_retention %>%
  select(threshold_numeric, true_reads_retained_pct, control_reads_retained_pct) %>%
  pivot_longer(
    cols = c(true_reads_retained_pct, control_reads_retained_pct),
    names_to = "sample_class",
    values_to = "reads_retained_pct"
  ) %>%
  mutate(sample_class = recode(
    sample_class,
    true_reads_retained_pct = "True samples",
    control_reads_retained_pct = "Negative controls"
  ))

p_retention <- ggplot(
  threshold_plot_data,
  aes(threshold_numeric, reads_retained_pct, colour = sample_class)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(0, 0.9, 0.1), labels = c("None", sprintf("%.1f", seq(0.1, 0.9, 0.1)))) +
  scale_colour_brewer(palette = "Dark2") +
  labs(x = "Decontam prevalence threshold", y = "Reads retained (%)", colour = NULL) +
  theme_bw(base_size = 10) + theme(legend.position = "bottom")

p_samples <- ggplot(
  threshold_retention,
  aes(threshold_numeric, samples_available_at_5876)
) +
  geom_line(colour = "#2c7fb8", linewidth = 0.8) +
  geom_point(colour = "#2c7fb8", size = 2) +
  geom_hline(yintercept = 50, linetype = 2, colour = "grey40") +
  scale_x_continuous(breaks = seq(0, 0.9, 0.1), labels = c("None", sprintf("%.1f", seq(0.1, 0.9, 0.1)))) +
  labs(x = "Decontam prevalence threshold", y = "Samples retained at 5,876 reads") +
  theme_bw(base_size = 10)

ggsave(file.path(fig_dir, "Figure1_decontam_read_and_sample_retention.png"),
       p_retention / p_samples, width = 9, height = 8, dpi = 600)
ggsave(file.path(fig_dir, "Figure1_decontam_read_and_sample_retention.tiff"),
       p_retention / p_samples, width = 9, height = 8, dpi = 600,
       compression = "lzw")

effect_plot_data <- permanova_all %>%
  filter(condition_axis == "decontam_threshold") %>%
  mutate(
    threshold_numeric = ifelse(condition_id == "None", 0, as.numeric(condition_id)),
    variable = recode(
      variable,
      subject_id = "Participant",
      site_group = "Anatomical site",
      swab_type2 = "Sampling system"
    ),
    distance = recode(
      distance,
      weighted_unifrac = "Weighted UniFrac",
      unweighted_unifrac = "Unweighted UniFrac",
      bray_curtis = "Bray-Curtis"
    )
  )

p_effect <- ggplot(
  effect_plot_data,
  aes(threshold_numeric, R2, colour = variable, group = variable)
) +
  geom_line(linewidth = 0.7) + geom_point(size = 1.7) +
  facet_wrap(~distance, ncol = 1) +
  scale_x_continuous(breaks = seq(0, 0.9, 0.1), labels = c("None", sprintf("%.1f", seq(0.1, 0.9, 0.1)))) +
  scale_colour_brewer(palette = "Dark2") +
  labs(x = "Decontam prevalence threshold", y = expression(PERMANOVA~R^2), colour = NULL) +
  theme_bw(base_size = 9) + theme(legend.position = "bottom")

ggsave(file.path(fig_dir, "Figure2_decontam_effect_size_stability.png"),
       p_effect, width = 8, height = 10, dpi = 600)
ggsave(file.path(fig_dir, "Figure2_decontam_effect_size_stability.tiff"),
       p_effect, width = 8, height = 10, dpi = 600, compression = "lzw")

p_removal <- sample_removal %>%
  filter(threshold == "0.5", control_status == "sample") %>%
  ggplot(aes(swab_type2, removed_pct, fill = swab_type2)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape = subject_id), position = position_jitter(width = 0.08), size = 2) +
  facet_wrap(~site_group) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Sampling system", y = "Reads removed by decontam 0.5 (%)", shape = "Participant") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1), legend.position = "bottom") +
  guides(fill = "none")

ggsave(file.path(fig_dir, "Figure3_sampling_system_read_removal.png"),
       p_removal, width = 9, height = 5, dpi = 600)
ggsave(file.path(fig_dir, "Figure3_sampling_system_read_removal.tiff"),
       p_removal, width = 9, height = 5, dpi = 600, compression = "lzw")

filter_plot <- filter_integrated %>%
  mutate(filter = factor(filter, levels = unname(filter_labels))) %>%
  ggplot(aes(reads_retained_vs_v05_pct, R2_swab_type2, colour = distance)) +
  geom_point(size = 2) +
  geom_text(aes(label = filter), size = 2.4, check_overlap = TRUE, hjust = -0.05) +
  facet_wrap(~distance, scales = "free_y") +
  labs(x = "Reads retained after filtering (%)", y = expression(Sampling-system~R^2), colour = NULL) +
  theme_bw(base_size = 9) + theme(legend.position = "none")

ggsave(file.path(fig_dir, "Figure4_filter_retention_and_system_effect.png"),
       filter_plot, width = 11, height = 4.5, dpi = 600)
ggsave(file.path(fig_dir, "Figure4_filter_retention_and_system_effect.tiff"),
       filter_plot, width = 11, height = 4.5, dpi = 600, compression = "lzw")

p_procrustes_threshold <- procrustes_results %>%
  filter(condition_axis == "decontam_threshold") %>%
  mutate(
    threshold_numeric = ifelse(condition_id == "None", 0, as.numeric(condition_id)),
    distance = recode(
      distance,
      weighted_unifrac = "Weighted UniFrac",
      unweighted_unifrac = "Unweighted UniFrac",
      bray_curtis = "Bray-Curtis"
    )
  ) %>%
  ggplot(aes(threshold_numeric, procrustes_correlation, colour = distance)) +
  geom_line() + geom_point(size = 2) +
  scale_x_continuous(breaks = seq(0, 0.9, 0.1), labels = c("None", sprintf("%.1f", seq(0.1, 0.9, 0.1)))) +
  scale_colour_brewer(palette = "Dark2") +
  labs(x = "Decontam prevalence threshold", y = "Procrustes correlation", colour = NULL) +
  theme_bw(base_size = 10) + theme(legend.position = "bottom")

p_procrustes_filter <- procrustes_results %>%
  filter(condition_axis == "filter_at_threshold_0.5") %>%
  mutate(
    condition_id = factor(condition_id, levels = names(filter_rarefied)),
    distance = recode(
      distance,
      weighted_unifrac = "Weighted UniFrac",
      unweighted_unifrac = "Unweighted UniFrac",
      bray_curtis = "Bray-Curtis"
    )
  ) %>%
  ggplot(aes(condition_id, procrustes_correlation, colour = distance, group = distance)) +
  geom_line() + geom_point(size = 2) +
  scale_colour_brewer(palette = "Dark2") +
  labs(x = "Filter at decontam threshold 0.5", y = "Procrustes correlation", colour = NULL) +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

p_procrustes <- p_procrustes_threshold / p_procrustes_filter +
  plot_layout(heights = c(1, 1.2))

ggsave(file.path(fig_dir, "Figure5_ordination_concordance.png"),
       p_procrustes, width = 10, height = 8, dpi = 600)
ggsave(file.path(fig_dir, "Figure5_ordination_concordance.tiff"),
       p_procrustes, width = 10, height = 8, dpi = 600, compression = "lzw")

p_shared <- aggregate_sharing_all %>%
  filter(taxonomic_level == "Genus") %>%
  mutate(
    analysis = recode(
      condition_axis,
      decontam_threshold = "Decontam threshold",
      filter_at_threshold_0.5 = "Filtering at threshold 0.5"
    )
  ) %>%
  ggplot(aes(condition_id, shared_all_systems_read_pct, group = 1)) +
  geom_line(colour = "#1b9e77") + geom_point(colour = "#1b9e77") +
  facet_wrap(~analysis, scales = "free_x") +
  labs(x = NULL, y = "Reads assigned to genera shared by all systems (%)") +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(fig_dir, "Figure6_shared_genus_read_fraction.png"),
       p_shared, width = 11, height = 5, dpi = 600)
ggsave(file.path(fig_dir, "Figure6_shared_genus_read_fraction.tiff"),
       p_shared, width = 11, height = 5, dpi = 600, compression = "lzw")

filter_display_levels <- names(filter_objects)
pair_class_labels <- c(
  within_participant_site_between_systems = "Between systems\nwithin participant-site",
  between_participants_matched_site_system = "Between participants\nmatched by site-system",
  between_sites_matched_participant_system = "Between sites\nmatched by participant-system"
)

p_fixed_change <- fixed_sample_distance_summary %>%
  filter(condition_id != "no_filter") %>%
  mutate(
    condition_id = factor(condition_id, levels = filter_display_levels),
    pair_class = recode(pair_class, !!!pair_class_labels),
    distance = recode(
      distance,
      weighted_unifrac = "Weighted UniFrac",
      unweighted_unifrac = "Unweighted UniFrac",
      bray_curtis = "Bray-Curtis"
    )
  ) %>%
  ggplot(aes(condition_id, median_percent_change, colour = pair_class,
             group = pair_class)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
  geom_line() + geom_point(size = 2) +
  facet_wrap(~distance, scales = "free_y") +
  scale_colour_brewer(palette = "Dark2") +
  labs(
    x = "Filter at decontam threshold 0.5",
    y = "Median distance change from no filter (%)",
    colour = NULL
  ) +
  theme_bw(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

ggsave(file.path(fig_dir, "Figure7_fixed50_matched_distance_change.png"),
       p_fixed_change, width = 12, height = 6, dpi = 600)
ggsave(file.path(fig_dir, "Figure7_fixed50_matched_distance_change.tiff"),
       p_fixed_change, width = 12, height = 6, dpi = 600, compression = "lzw")

p_signal_ratio <- distance_signal_ratios %>%
  select(condition_id, distance,
         participant_to_system_ratio, site_to_system_ratio) %>%
  pivot_longer(
    cols = c(participant_to_system_ratio, site_to_system_ratio),
    names_to = "signal",
    values_to = "signal_to_system_ratio"
  ) %>%
  mutate(
    condition_id = factor(condition_id, levels = filter_display_levels),
    signal = recode(
      signal,
      participant_to_system_ratio = "Participant/system-distance ratio",
      site_to_system_ratio = "Site/system-distance ratio"
    ),
    distance = recode(
      distance,
      weighted_unifrac = "Weighted UniFrac",
      unweighted_unifrac = "Unweighted UniFrac",
      bray_curtis = "Bray-Curtis"
    )
  ) %>%
  ggplot(aes(condition_id, signal_to_system_ratio, colour = signal,
             group = signal)) +
  geom_line() + geom_point(size = 2) +
  facet_wrap(~distance, scales = "free_y") +
  scale_colour_brewer(palette = "Set1") +
  labs(
    x = "Filter at decontam threshold 0.5",
    y = "Biological-signal / within-block system-distance ratio",
    colour = NULL
  ) +
  theme_bw(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

ggsave(file.path(fig_dir, "Figure8_fixed50_signal_preservation_ratio.png"),
       p_signal_ratio, width = 12, height = 6, dpi = 600)
ggsave(file.path(fig_dir, "Figure8_fixed50_signal_preservation_ratio.tiff"),
       p_signal_ratio, width = 12, height = 6, dpi = 600, compression = "lzw")
```

# Reproducibility record

```{r reproducibility}
final_checks <- bind_rows(
  validation %>% select(check, passed),
  tibble(
    check = c(
      "all PERMANOVA tests record 9999 permutations",
      "all PERMDISP tests record 9999 permutations",
      "all PROTEST permutation runs record 9999 permutations",
      "no tree transformation function used in analysis",
      "threshold axis contains None plus 0.1-0.9",
      "filter axis contains all nine prespecified filters",
      "fixed-sample filtering comparison retains the same 50 samples",
      "fixed-sample matched-distance units have expected block counts"
    ),
    passed = c(
      all(permanova_all$permutations == permutations_n),
      all(permdisp_all$permutations == permutations_n),
      all(procrustes_results$permutations == permutations_n),
      identical(phy_tree(phy_pre), original_tree),
      setequal(names(threshold_objects), threshold_labels),
      setequal(names(filter_objects), filter_ids),
      all(vapply(filter_distance_objects, nsamples, integer(1)) == 50L),
      all(
        fixed_sample_distance_units %>%
          distinct(condition_id, distance, pair_class, unit_id) %>%
          count(condition_id, distance, pair_class) %>%
          mutate(expected = case_when(
            pair_class == "within_participant_site_between_systems" ~ 10L,
            pair_class == "between_participants_matched_site_system" ~ 10L,
            pair_class == "between_sites_matched_participant_system" ~ 25L
          )) %>%
          transmute(ok = n == expected) %>%
          pull(ok)
      )
    )
  )
)
write.csv(final_checks, file.path(tab_dir, "Table16_final_sanity_checks.csv"), row.names = FALSE)
if (!all(final_checks$passed)) stop("One or more integrated final checks failed")

writeLines(
  c(
    "Integrated decontam, filtering, and sampling-system sensitivity analysis",
    paste("Generated:", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")),
    "Primary question: stability of the five-system comparison",
    "Input: Phyloseq/phy_F270R240_260812_v3.rds",
    "Saved canonical comparison: Phyloseq/phy_F270R240_260812_v4.rds",
    "Existing phylogenetic tree used without modification",
    "Threshold axis: no decontam and prevalence thresholds 0.1-0.9 with singleton removal",
    "Filter axis: nine filters at decontam threshold 0.5",
    "Fixed-sample distance validation: the same 50 samples under every filter",
    "Paired units: participant-site, site-system, and participant-system blocks",
    "Rarefaction depth: 5876",
    "Permutation tests: 9999",
    "Random seed: 42",
    "Thresholds causing sample loss are boundary conditions and are not primary comparable analyses."
  ),
  file.path(out_dir, "README_results.txt")
)
capture.output(sessionInfo(), file = file.path(out_dir, "sessionInfo.txt"))
final_checks
```

# Interpretation guardrails

- The primary topic is the comparison of the five integrated sampling systems.
- Decontamination and filtering are evaluated as sources of analytical sensitivity,
  not as a search for a preferred p-value.
- Non-significant system tests do not establish equivalence or interchangeability.
- Thresholds or filters that remove samples at 5,876 reads change the analysis set
  and are interpreted as boundary conditions.
- Total DNA and sequencing reads are not direct measures of absolute bacterial load.
- Aggregate sharing is reported together with participant-by-site matched-block
  sharing to avoid overinterpreting pooled intersections.
