---
title: "Decontam threshold × rare-feature filtering benchmark"
author: "DT Swab Study"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
project_library <- file.path(getwd(), "R_library")
if (dir.exists(project_library)) .libPaths(c(project_library, .libPaths()))
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Objective and prespecified design

This benchmark tests whether the relative ordering of participant, anatomical-site,
and sampling-system effects remains stable across prevalence-based decontamination
thresholds 0.1--0.9 and three prespecified rare-feature filters. It is a sensitivity
analysis, not a search for a combination producing a preferred p-value.

The response is community dissimilarity (weighted UniFrac, unweighted UniFrac, or
Bray--Curtis). Explanatory variables are tested in separate univariable PERMANOVA
models. Participant identity uses unrestricted permutations; anatomical site and
sampling system use permutations restricted within participant. Sampling-system
PERMDISP is evaluated with the same participant restriction. All tests use 9,999
permutations and a fixed rarefaction target of 5,876 reads.

```{r packages}
suppressPackageStartupMessages({
  library(phyloseq)
  library(ape)
  library(decontam)
  library(vegan)
  library(permute)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(patchwork)
  library(RColorBrewer)
})
set.seed(42)

out_dir <- file.path(getwd(), "Decontam_filter_benchmark_output")
fig_dir <- file.path(out_dir, "Figures")
tab_dir <- file.path(out_dir, "Tables")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(tab_dir, recursive = TRUE, showWarnings = FALSE)

thresholds <- seq(0.1, 0.9, by = 0.1)
filter_ids <- c("no_filter", "singleton", "count_le_10")
filter_labels <- c(
  no_filter = "No additional filter",
  singleton = "Singleton removal",
  count_le_10 = "Remove total count <=10"
)
distance_methods <- c(
  weighted_unifrac = "wunifrac",
  unweighted_unifrac = "unifrac",
  bray_curtis = "bray"
)
target_depth <- 5876L

# phyloseq's fast UniFrac implementation assumes a strictly bifurcating tree.
# Feature pruning leaves a small number of polytomies in these objects. Resolve
# them deterministically with zero-length branches; this changes no original
# branch length or tip identity and prevents the fast-UniFrac matrix warning.
prepare_unifrac_tree <- function(ps) {
  tree <- phy_tree(ps)
  if (!ape::is.binary(tree)) {
    tree <- ape::multi2di(tree, random = FALSE)
    phy_tree(ps) <- tree
  }
  stopifnot(ape::is.binary(phy_tree(ps)))
  ps
}
```

# Reconstruct all analysis objects

```{r reconstruct-objects}
otu_matrix <- function(ps) {
  otu <- as(otu_table(ps), "matrix")
  if (!taxa_are_rows(ps)) otu <- t(otu)
  otu
}

phy_pre <- readRDS("Phyloseq/phy_F270R240_260812_v3.rds")

# Restore the control labels from the original metadata by exact sample ID.
# In the saved v3 object, the five controls have NA control_status because the
# original preprocessing converted these missing values to "control" only after
# creating v3. The metadata records them explicitly as "negative-control".
metadata_source <- read.delim(
  "metadata/DT_metadata.tsv",
  check.names = FALSE,
  stringsAsFactors = FALSE
)
rownames(metadata_source) <- metadata_source$sample_id
stopifnot(all(sample_names(phy_pre) %in% rownames(metadata_source)))
metadata_source <- metadata_source[sample_names(phy_pre), , drop = FALSE]
metadata_source$control_status <- ifelse(
  metadata_source$control_status == "negative-control",
  "control",
  metadata_source$control_status
)
stopifnot(
  sum(metadata_source$control_status == "sample") == 50L,
  sum(metadata_source$control_status == "control") == 5L
)
sample_data(phy_pre) <- sample_data(metadata_source)
sample_data(phy_pre)$is.neg <- sample_data(phy_pre)$control_status == "control"

# Follow 01_Preprocessing.Rmd exactly: rerun prevalence-based decontam at each
# threshold and use the returned `contaminant` classification.
decontam_results <- setNames(
  lapply(thresholds, function(threshold) {
    isContaminant(
      phy_pre,
      method = "prevalence",
      neg = "is.neg",
      threshold = threshold
    )
  }),
  sprintf("%0.1f", thresholds)
)

apply_rare_filter <- function(ps, filter_id) {
  totals <- taxa_sums(ps)
  keep <- switch(
    filter_id,
    no_filter = totals > 0,
    singleton = totals > 1,
    count_le_10 = totals > 10,
    stop("Unknown filter: ", filter_id)
  )
  prune_taxa(keep, ps)
}

make_combination <- function(threshold, filter_id) {
  contam_result <- decontam_results[[sprintf("%0.1f", threshold)]]
  clean <- prune_taxa(!contam_result$contaminant, phy_pre)
  true <- subset_samples(clean, control_status == "sample")
  true <- prune_taxa(taxa_sums(true) > 0, true)
  apply_rare_filter(true, filter_id)
}

combination_grid <- tidyr::expand_grid(
  threshold = thresholds,
  filter_id = filter_ids
) %>%
  mutate(
    combination_id = sprintf("decontam_%0.1f__%s", threshold, filter_id),
    filter = unname(filter_labels[filter_id])
  )

analysis_objects <- setNames(
  lapply(seq_len(nrow(combination_grid)), function(i) {
    make_combination(combination_grid$threshold[i], combination_grid$filter_id[i])
  }),
  combination_grid$combination_id
)

rarefy_object <- function(ps) {
  keep <- sample_names(ps)[sample_sums(ps) >= target_depth]
  if (length(keep) < 2L) return(NULL)
  ps <- prune_samples(keep, ps)
  ps <- prune_taxa(taxa_sums(ps) > 0, ps)
  set.seed(42)
  rarefy_even_depth(ps, sample.size = target_depth, rngseed = 42,
                    replace = FALSE, trimOTUs = TRUE, verbose = FALSE)
}
rarefied_objects <- lapply(analysis_objects, rarefy_object)
```

# Data retention and sanity checks

```{r retention}
pre_true <- subset_samples(phy_pre, control_status == "sample")
pre_true <- prune_taxa(taxa_sums(pre_true) > 0, pre_true)
baseline_id <- "decontam_0.5__no_filter"
baseline <- analysis_objects[[baseline_id]]

retention <- bind_rows(lapply(seq_len(nrow(combination_grid)), function(i) {
  id <- combination_grid$combination_id[i]
  ps <- analysis_objects[[id]]
  rar <- rarefied_objects[[id]]
  tibble(
    threshold = combination_grid$threshold[i],
    filter_id = combination_grid$filter_id[i],
    filter = combination_grid$filter[i],
    combination_id = id,
    retained_asvs = ntaxa(ps),
    retained_reads = sum(sample_sums(ps)),
    retained_samples = nsamples(ps),
    samples_below_5876 = sum(sample_sums(ps) < target_depth),
    samples_rarefied_5876 = if (is.null(rar)) 0L else nsamples(rar),
    minimum_depth = min(sample_sums(ps)),
    read_retained_vs_predecontam_pct = 100 * sum(sample_sums(ps)) / sum(sample_sums(pre_true))
  )
}))
write.csv(retention, file.path(tab_dir, "Table1_retention_27_combinations.csv"), row.names = FALSE)

# Threshold 0.5/no-filter must reproduce the saved post-decontam v4 true samples.
phy_v4 <- readRDS("Phyloseq/phy_F270R240_260812_v4.rds")
phy_v4_true <- subset_samples(phy_v4, control_status == "sample")
phy_v4_true <- prune_taxa(taxa_sums(phy_v4_true) > 0, phy_v4_true)
sanity_checks <- tibble(
  check = c(
    "27 combinations reconstructed",
    "threshold 0.5 ASV count matches v4",
    "threshold 0.5 read count matches v4",
    "all combinations retain true-sample metadata"
  ),
  passed = c(
    length(analysis_objects) == 27L,
    ntaxa(baseline) == ntaxa(phy_v4_true),
    sum(sample_sums(baseline)) == sum(sample_sums(phy_v4_true)),
    all(vapply(analysis_objects, nsamples, integer(1)) == 50L)
  ),
  observed = c(
    as.character(length(analysis_objects)),
    paste(ntaxa(baseline), ntaxa(phy_v4_true), sep = " vs "),
    paste(sum(sample_sums(baseline)), sum(sample_sums(phy_v4_true)), sep = " vs "),
    paste(range(vapply(analysis_objects, nsamples, integer(1))), collapse = "-")
  )
)
write.csv(sanity_checks, file.path(tab_dir, "validation_sanity_checks.csv"), row.names = FALSE)
if (!all(sanity_checks$passed)) stop("Benchmark reconstruction sanity check failed")
sanity_checks
```

# PERMANOVA and PERMDISP

```{r permanova-permdisp}
run_permanova <- function(ps, distance_name, method, variable) {
  if (method %in% c("wunifrac", "unifrac")) ps <- prepare_unifrac_tree(ps)
  dist_obj <- phyloseq::distance(ps, method = method)
  md <- data.frame(sample_data(ps), check.names = FALSE)
  formula_obj <- as.formula(paste("dist_obj ~", variable))
  blocked <- variable != "subject_id"
  set.seed(42)
  fit <- if (blocked) {
    adonis2(formula_obj, data = md, permutations = 9999,
            strata = md$subject_id)
  } else {
    adonis2(formula_obj, data = md, permutations = 9999)
  }
  tibble(
    distance_id = distance_name,
    variable = variable,
    n_samples = nrow(md),
    df = fit$Df[1],
    pseudo_F = fit$F[1],
    R2 = fit$R2[1],
    p_value = fit$`Pr(>F)`[1],
    permutation_scheme = ifelse(blocked, "within participant", "unrestricted")
  )
}

run_permdisp <- function(ps, distance_name, method) {
  if (method %in% c("wunifrac", "unifrac")) ps <- prepare_unifrac_tree(ps)
  dist_obj <- phyloseq::distance(ps, method = method)
  md <- data.frame(sample_data(ps), check.names = FALSE)
  dispersion <- betadisper(dist_obj, md$swab_type2, type = "median", bias.adjust = TRUE)
  design <- permute::how(nperm = 9999, blocks = md$subject_id)
  set.seed(42)
  fit <- permutest(dispersion, permutations = design)
  tibble(
    distance_id = distance_name,
    permdisp_F = fit$tab$F[1],
    permdisp_p = fit$tab$`Pr(>F)`[1]
  )
}

permanova_results <- bind_rows(lapply(seq_len(nrow(combination_grid)), function(i) {
  id <- combination_grid$combination_id[i]
  ps <- rarefied_objects[[id]]
  bind_rows(lapply(names(distance_methods), function(distance_name) {
    bind_rows(lapply(c("subject_id", "site_group", "swab_type2"), function(variable) {
      run_permanova(ps, distance_name, distance_methods[[distance_name]], variable)
    }))
  })) %>%
    mutate(
      threshold = combination_grid$threshold[i],
      filter_id = combination_grid$filter_id[i],
      filter = combination_grid$filter[i],
      combination_id = id,
      .before = 1
    )
}))

permdisp_results <- bind_rows(lapply(seq_len(nrow(combination_grid)), function(i) {
  id <- combination_grid$combination_id[i]
  ps <- rarefied_objects[[id]]
  bind_rows(lapply(names(distance_methods), function(distance_name) {
    run_permdisp(ps, distance_name, distance_methods[[distance_name]])
  })) %>%
    mutate(
      threshold = combination_grid$threshold[i],
      filter_id = combination_grid$filter_id[i],
      filter = combination_grid$filter[i],
      combination_id = id,
      .before = 1
    )
}))

write.csv(permanova_results, file.path(tab_dir, "Table2_PERMANOVA_27_combinations.csv"), row.names = FALSE)
write.csv(permdisp_results, file.path(tab_dir, "Table3_PERMDISP_27_combinations.csv"), row.names = FALSE)
```

# Composition and shared-feature stability

```{r composition-shared}
composition_similarity <- function(reference, candidate) {
  shared_samples <- intersect(sample_names(reference), sample_names(candidate))
  ref <- prune_samples(shared_samples, reference)
  cand <- prune_samples(shared_samples, candidate)
  all_taxa <- union(taxa_names(ref), taxa_names(cand))
  make_aligned <- function(ps) {
    otu <- otu_matrix(ps)
    aligned <- matrix(0, nrow = length(all_taxa), ncol = length(shared_samples),
                      dimnames = list(all_taxa, shared_samples))
    aligned[rownames(otu), colnames(otu)] <- otu
    sweep(aligned, 2, colSums(aligned), "/")
  }
  ref_rel <- make_aligned(ref)
  cand_rel <- make_aligned(cand)
  tibble(
    sample_name = shared_samples,
    similarity = vapply(shared_samples, function(s) {
      1 - sum(abs(ref_rel[, s] - cand_rel[, s])) / 2
    }, numeric(1))
  )
}

shared_summary <- function(ps, rank_name) {
  work <- if (rank_name == "ASV") ps else tax_glom(ps, taxrank = "Genus", NArm = FALSE)
  otu <- otu_matrix(work)
  md <- data.frame(sample_data(work), check.names = FALSE)
  systems <- sort(unique(as.character(md$swab_type2)))
  present <- sapply(systems, function(system) {
    rowSums(otu[, rownames(md)[md$swab_type2 == system], drop = FALSE]) > 0
  })
  shared <- rownames(otu)[rowSums(present) == length(systems)]
  exclusive <- rownames(otu)[rowSums(present) == 1L]
  tibble(
    taxonomic_level = rank_name,
    total_features = nrow(otu),
    shared_features = length(shared),
    shared_read_pct = 100 * sum(otu[shared, , drop = FALSE]) / sum(otu),
    exclusive_features = length(exclusive),
    exclusive_read_pct = 100 * sum(otu[exclusive, , drop = FALSE]) / sum(otu)
  )
}

composition_results <- bind_rows(lapply(seq_len(nrow(combination_grid)), function(i) {
  id <- combination_grid$combination_id[i]
  composition_similarity(baseline, analysis_objects[[id]]) %>%
    summarise(
      mean_similarity = mean(similarity),
      median_similarity = median(similarity),
      minimum_similarity = min(similarity)
    ) %>%
    mutate(
      threshold = combination_grid$threshold[i],
      filter_id = combination_grid$filter_id[i],
      filter = combination_grid$filter[i],
      combination_id = id,
      .before = 1
    )
}))

shared_results <- bind_rows(lapply(seq_len(nrow(combination_grid)), function(i) {
  id <- combination_grid$combination_id[i]
  bind_rows(
    shared_summary(analysis_objects[[id]], "ASV"),
    shared_summary(analysis_objects[[id]], "Genus")
  ) %>%
    mutate(
      threshold = combination_grid$threshold[i],
      filter_id = combination_grid$filter_id[i],
      filter = combination_grid$filter[i],
      combination_id = id,
      .before = 1
    )
}))

write.csv(composition_results, file.path(tab_dir, "Table4_composition_similarity.csv"), row.names = FALSE)
write.csv(shared_results, file.path(tab_dir, "Table5_shared_exclusive_features.csv"), row.names = FALSE)
```

# Integrated benchmark and figures

```{r integrated-results, fig.width=11, fig.height=8}
effect_wide <- permanova_results %>%
  select(threshold, filter_id, filter, combination_id, distance_id, variable, R2, p_value) %>%
  pivot_wider(names_from = variable, values_from = c(R2, p_value))

integrated <- retention %>%
  left_join(effect_wide, by = c("threshold", "filter_id", "filter", "combination_id")) %>%
  left_join(permdisp_results, by = c("threshold", "filter_id", "filter", "combination_id", "distance_id")) %>%
  left_join(composition_results, by = c("threshold", "filter_id", "filter", "combination_id")) %>%
  left_join(
    shared_results %>%
      filter(taxonomic_level == "Genus") %>%
      select(threshold, filter_id, combination_id, shared_genus_read_pct = shared_read_pct,
             exclusive_genus_read_pct = exclusive_read_pct),
    by = c("threshold", "filter_id", "combination_id")
  ) %>%
  mutate(
    participant_to_system_R2_ratio = R2_subject_id / R2_swab_type2,
    site_to_system_R2_ratio = R2_site_group / R2_swab_type2,
    effect_order_supported = R2_subject_id > R2_site_group & R2_site_group > R2_swab_type2,
    system_significant = p_value_swab_type2 < 0.05,
    dispersion_significant = permdisp_p < 0.05
  )
write.csv(integrated, file.path(tab_dir, "Table6_integrated_benchmark.csv"), row.names = FALSE)

p_retention <- ggplot(retention, aes(threshold, read_retained_vs_predecontam_pct,
                                     colour = filter, group = filter)) +
  geom_line() + geom_point() +
  facet_wrap(~filter, nrow = 1) +
  labs(x = "Decontam threshold", y = "True-sample reads retained (%)",
       colour = "Rare-feature filter") +
  theme_bw(base_size = 9) + theme(legend.position = "none")

p_effect <- permanova_results %>%
  mutate(variable = recode(variable, subject_id = "Participant",
                           site_group = "Anatomical site", swab_type2 = "Sampling system")) %>%
  ggplot(aes(threshold, R2, colour = variable, group = variable)) +
  geom_line() + geom_point(size = 1.4) +
  facet_grid(distance_id ~ filter) +
  scale_colour_brewer(palette = "Dark2") +
  labs(x = "Decontam threshold", y = expression(PERMANOVA~R^2), colour = NULL) +
  theme_bw(base_size = 8) + theme(legend.position = "bottom")

p_diagnostics <- integrated %>%
  ggplot(aes(threshold, permdisp_p, colour = filter, group = filter)) +
  geom_hline(yintercept = 0.05, linetype = 2, colour = "red") +
  geom_line() + geom_point() +
  facet_wrap(~distance_id) +
  labs(x = "Decontam threshold", y = "Sampling-system PERMDISP p-value", colour = NULL) +
  theme_bw(base_size = 9) + theme(legend.position = "bottom")

ggsave(file.path(fig_dir, "Figure1_retention_and_effect_size_benchmark.png"),
       p_retention / p_effect, width = 11, height = 10, dpi = 600)
ggsave(file.path(fig_dir, "Figure1_retention_and_effect_size_benchmark.tiff"),
       p_retention / p_effect, width = 11, height = 10, dpi = 600,
       compression = "lzw")
ggsave(file.path(fig_dir, "Figure2_PERMDISP_benchmark.png"),
       p_diagnostics, width = 9, height = 5, dpi = 600)
ggsave(file.path(fig_dir, "Figure2_PERMDISP_benchmark.tiff"),
       p_diagnostics, width = 9, height = 5, dpi = 600, compression = "lzw")

benchmark_summary <- integrated %>%
  group_by(distance_id) %>%
  summarise(
    combinations = n(),
    participant_R2_min = min(R2_subject_id),
    participant_R2_max = max(R2_subject_id),
    site_R2_min = min(R2_site_group),
    site_R2_max = max(R2_site_group),
    system_R2_min = min(R2_swab_type2),
    system_R2_max = max(R2_swab_type2),
    system_p_min = min(p_value_swab_type2),
    system_p_max = max(p_value_swab_type2),
    permdisp_p_min = min(permdisp_p),
    effect_order_supported_n = sum(effect_order_supported),
    system_significant_n = sum(system_significant),
    dispersion_significant_n = sum(dispersion_significant),
    .groups = "drop"
  )
write.csv(benchmark_summary, file.path(tab_dir, "Table7_benchmark_summary.csv"), row.names = FALSE)
benchmark_summary
```

# Interpretation guardrails

- Threshold and filtering combinations are prespecified sensitivity conditions;
  none should be selected because it produces a desirable p-value.
- A non-significant sampling-system test does not establish equivalence or
  interchangeability, particularly with five participants.
- Conditions that lose samples at the fixed rarefaction depth are not directly
  comparable to conditions retaining all 50 samples without acknowledging the
  changed analysis set.
- Aggregate shared-feature results do not replace participant-by-site matched-block
  concordance analysis.

```{r reproducibility}
writeLines(capture.output(sessionInfo()), file.path(out_dir, "sessionInfo.txt"))
writeLines(
  c(
    paste("Generated:", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")),
    "Input: Phyloseq/phy_F270R240_260812_v3.rds",
    "Design: decontam thresholds 0.1-0.9 x 3 rare-feature filters",
    "Rarefaction depth: 5876",
    "Permutations: 9999",
    "Baseline for composition similarity: threshold 0.5, no additional filter"
  ),
  file.path(out_dir, "README_results.txt")
)
sessionInfo()
```
